---
title: "In class Exercise 06"
author: "Gautamgovan"
date: "Oct 4, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 4 
    number-sections: true
    
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Loading neccessary packages
```{r}
pacman::p_load(sf, tmap, tidyverse, rvest, knitr)
```

## Importing Data 

### Loading Master Plan2019 Subzone Boundary NoSea KML

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

```{r}
mpsz
```
```{r}
mpsz <- st_transform(mpsz, 
                              crs = 3414)
mpsz
```
### Busstop
```{r}
BusStop <- st_read(dsn = "data/geospatial/BusStopLocation_Aug2025",layer="BusStop")
```
### O-D Bus Trips

```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202508.csv")
```

Making sure both mpsz and BusStop are following same crs code 

```{r}
BusStop <- st_transform(BusStop, 
                              crs = 3414)
BusStop
```
### PLotting map 

```{r}
tm_shape(mpsz)+ tm_polygons()+ tm_shape(BusStop)+ tm_dots()
```
## Extracting bus stop within Singapore 

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz,
  join = st_within,
  left = FALSE
)
```

Check again

```{r}
tm_shape(mpsz)+ tm_polygons()+ tm_shape(BusStop_in_SG)+ tm_dots()
```
We can find that all bus stops are within Singapore 

## Derving Analytical Hexagon 

```{r}
hexagon <- st_make_grid(mpsz,
  cellsize = 700,
  what = "polygon",
  square = FALSE
) %>%
  st_sf()
```

In Hands on exercise check whether cell size= 700 really means 375 meters in hexagon as required by Hands-on- Exercise 

## Filtering hexagons with bus stops

```{r}
hexagon$n_collisions = 
  lengths(st_intersects(
    hexagon, BusStop_in_SG
  ))

hexagon <- filter(hexagon, n_collisions > 0)
```

## Assigning ids to each hexagon 

```{r}
hexagon <- hexagon %>%
  select(-n_collisions)

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(nrow(hexagon))
) %>%
  as.factor()

kable(head(hexagon))
```
## Preparing Trip Generation Data 
### Cleaning the data 
Remove unnecessary column and aggregate 

```{r}
trips <- odbus %>%
  select(c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%
  rename(HOUR_OF_DAY = TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS)

trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N)

kable(head(trips))
```
### Combine 
```{r}
bs_hex <- st_intersection(
  BusStop, hexagon
) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID))

kable(head(bs_hex))

```
## Adding HEX_ID into bus trips data 

```{r}
trips <- inner_join(trips, bs_hex)
kable(head(trips))
```











