---
title: "Hands On Exercise 6"
author: "Gautamgovan"
date: "Oct 2, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 4 
    number-sections: true
    
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true 
---

# Geographical Segmentation with Spatially Constrained Clustering Techniques

## Overview 
In this hands-on exercise, we will gain hands-on experience on how to delineate homogeneous regions by using geographically referenced multivariate data. There are two major analyses, namely:

- **Hierarchical cluster analysis**; and  
- **Spatially constrained cluster analysis**.

## Getting Started 

### The analytical question

In geobusiness and spatial policy, it is a common practice to delineate the market or planning area into homogeneous regions by using multivariate data. In this hands-on exercise, we are interested to delineate Shan State, Myanmar into homogeneous regions by using multiple Information and Communication technology (ICT) measures, namely: Radio, Television, Land line phone, Mobile phone, Computer, and Internet at home.

## The data
Two data sets will be used in this study. They are:

- **Myanmar Township Boundary Data**  
  (i.e. *myanmar_township_boundaries*): This is a GIS data in ESRI shapefile format. It consists of township boundary information of Myanmar. The spatial data are captured in polygon features.

- **Shan-ICT.csv**: This is an extract of [The 2014 Myanmar Population and Housing Census Myanmar](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level.

### Installing and loading R packages

Before we get started, it is important for us to install the necessary R packages into R and launch these R packages into R environment.  

The R packages needed for this exercise are as follows:

- **Spatial data handling**
  - *sf*, *rgdal*, and *spdep*
- **Attribute data handling**
  - *tidyverse*, especially *readr*, *ggplot2*, and *dplyr*
- **Choropleth mapping**
  - *tmap*
- **Multivariate data visualisation and analysis**
  - *corrplot*, *ggpubr*, and *heatmaply*
- **Cluster analysis**
  - *cluster*
  - *ClustGeo*

The code chunks below install and launch these R packages into R environment.

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```
##  Data Import and Prepatation

###  Importing geospatial data into R environment

In this section, we will import Myanmar Township Boundary GIS data and its associated attrbiute table into R environment.

The Myanmar Township Boundary GIS data is in ESRI shapefile format. It will be imported into R environment by using the st_read() function of sf.

The code chunks used are shown below:

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

The imported township boundary object is called shan_sf. It is saved in simple feature data.frame format. We can view the content of the newly created shan_sf simple features data.frame by using the code chunk below.

```{r}
shan_sf
```

```{r}
glimpse(shan_sf)
```
### Importing aspatial data into R environment

The csv file will be import using read_csv function of readr package.

The code chunks used are shown below:

```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```
```{r}
summary(ict)
```
There are a total of eleven fields and 55 observation in the tibble data.frame.

### Derive new variables using dplyr package

The unit of measurement of the values are number of household. Using these values directly will be bias by the underlying total number of households. In general, the townships with relatively higher total number of households will also have higher number of households owning radio, TV, etc.

In order to overcome this problem, we will derive the penetration rate of each ICT variable by using the code chunk below.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

Let us review the summary statistics of the newly derived penetration rates using the code chunk below.

```{r}

summary(ict_derived)
```

Notice that six new fields have been added into the data.frame. They are RADIO_PR, TV_PR, LLPHONE_PR, MPHONE_PR, COMPUTER_PR, and INTERNET_PR.

## Exploratory Data Analysis (EDA)

### EDA using statistical graphics

We can plot the distribution of the variables (i.e. Number of households with radio) by using appropriate Exploratory Data Analysis (EDA) as shown in the code chunk below.

Histogram is useful to identify the overall distribution of the data values (i.e. left skew, right skew or normal distribution)

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```
Boxplot is useful to detect if there are outliers

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Next, we will also plotting the distribution of the newly derived variables (i.e. Radio penetration rate) by using the code chunk below.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```
```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

The code chunks below are used to create the data visualisation. They consist of two main parts

multiple histograms are plotted to reveal the distribution of the selected variables in the ict_derived data.frame.

First, we will create the individual histograms using the code chunk below.
```{r}
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```


Next, the ggarrange() function of ggpubr package is used to group these histograms together.

```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

