---
title: "Hands On Exercise 8"
author: "Gautamgovan"
date: "October 23, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 4
    number-sections: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---
## Overview
Predictive modelling uses statistical learning or machine learning techniques to predict outcomes. By and large, the event one wants to predict is in the future. However, a set of known outcome and predictors (also known as variables) will be used to calibrate the predictive models.

Geospatial predictive modelling is conceptually rooted in the principle that the occurrences of events being modeled are limited in distribution. When geographically referenced data are used, occurrences of events are neither uniform nor random in distribution over space. There are geospatial factors (infrastructure, sociocultural, topographic, etc.) that constrain and influence where the locations of events occur. Geospatial predictive modeling attempts to describe those constraints and influences by spatially correlating occurrences of historical geospatial locations with environmental factors that represent those constraints and influences.

In this hands on exercise, we will learn how to build predictive model by using geographical random forest method. By the end of this hands-on exercise, you will acquire the skills of:

- preparing training and test data sets by using appropriate data sampling methods,
- calibrating predictive models by using both geospatial statistical learning and machine learning methods,
- comparing and selecting the best model for predicting the future outcome,
- predicting the future outcomes by using the best model calibrated.

## The Data
### Aspatial Dataset

**HDB Resale Data**  
A list of HDB resale transacted prices in Singapore from January 2017 onwards.  
**Format:** CSV  
**Source:** [Data.gov.sg](https://data.gov.sg)

---

### Geospatial Dataset

**MP14A_SUBZONE_WEB_PL**  
Polygon feature data providing information on URA 2014 Master Plan Planning Subzone boundaries.  
**Format:** ESRI Shapefile  
**Source:** [Data.gov.sg](https://data.gov.sg)

---

### Locational Factors with Geographic Coordinates

**Downloaded from [Data.gov.sg](https://data.gov.sg):**  
- **Eldercare data:** List of eldercare centres in Singapore (Shapefile format).  
- **Hawker Centre data:** List of hawker centres in Singapore (GeoJSON format).  
- **Parks data:** List of parks in Singapore (GeoJSON format).  
- **Supermarket data:** List of supermarkets in Singapore (GeoJSON format).  
- **CHAS clinics data:** List of CHAS clinics in Singapore (GeoJSON format).  
- **Childcare service data:** List of childcare services in Singapore (GeoJSON format).  
- **Kindergartens data:** List of kindergartens in Singapore (GeoJSON format).

**Downloaded from [Datamall.lta.gov.sg](https://datamall.lta.gov.sg):**  
- **MRT data:** List of MRT/LRT stations in Singapore, including station names and codes.  
  **Format:** Shapefile  
- **Bus stops data:** List of bus stops in Singapore.  
  **Format:** Shapefile

---

### Locational Factors without Geographic Coordinates

**Downloaded from [Data.gov.sg](https://data.gov.sg):**  
- **Primary school data:** Extracted from the list of general information on schools.  
  **Format:** CSV

**Retrieved/Scraped from Other Sources:**  
- **CBD coordinates:** Obtained from Google.  
- **Shopping malls data:** List of shopping malls in Singapore, obtained from Wikipedia.  
- **Good primary schools:** List of primary schools ranked by popularity, available from Local Salary Forum.

## Installing and Loading R packages

This code chunk performs 3 tasks:

- A list called packages will be created and will consists of all the R packages required to accomplish this exercise.
- Check if R packages on package have been installed in R and if not, they will be installed.
- After all the R packages have been installed, they will be loaded.

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, Metrics, tidyverse)
```
## Preparing Data
### Reading data file to rds
Reading the input data sets. It is in simple feature data frame.
```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```

```{r}
mdata
```


### Data Sampling
The entire data are split into training and test data sets with 65% and 35% respectively by using initial_split() of rsample package. rsample is one of the package of tigymodels.

```{r}
set.seed(1234)
resale_split <- initial_split(mdata, 
                              prop = 6.5/10,)
train_data <- training(resale_split)
test_data <- testing(resale_split)
```

```{r}
write_rds(train_data, "data/model/train_data.rds")
write_rds(test_data, "data/model/test_data.rds")
```

## Computing Correlation Matrix

Before loading the predictors into a predictive model, it is always a good practice to use correlation matrix to examine if there is sign of multicolinearity.

```{r}
mdata_nogeo <- mdata %>%
  st_drop_geometry()
corrplot::corrplot(cor(mdata_nogeo[, 2:17]), 
                   diag = FALSE, 
                   order = "AOE",
                   tl.pos = "td", 
                   tl.cex = 0.5, 
                   method = "number", 
                   type = "upper")
```
##  Retriving the Stored Data

```{r}
train_data <- read_rds("data/model/train_data.rds")
test_data <- read_rds("data/model/test_data.rds")
```

## Building a non-spatial multiple linear regression

```{r}
price_mlr <- lm(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                data=train_data)
summary(price_mlr)
```
```{r}
write_rds(price_mlr, "data/model/price_mlr.rds" ) 
```

##  gwr predictive method
In this section, we will learn how to calibrate a model to predict HDB resale price by using geographically weighted regression method of GWmodel package.

```{r}
#| eval: false
bw_adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
    WITHIN_1KM_PRISCH,
  data=train_data,
  approach="CV",
  kernel="gaussian",
  adaptive=TRUE,
  longlat=FALSE)

write_rds(bw_adaptive, "data/model/bw_adaptive.rds")
```
## Constructing the adaptive bandwidth gwr model

First, let us call the save bandwidth by using the code chunk below.

```{r}
#| include: false
bw_adaptive <- read_rds(
  "data/model/bw_adaptive.rds")
```

Now, we can go ahead to calibrate the gwr-based hedonic pricing model by using adaptive bandwidth and Gaussian kernel as shown in the code chunk below.

```{r}
#| eval: false
gwr_adaptive <- gwr.basic(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  data=train_data,
  bw=bw_adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE,
  longlat = FALSE)

write_rds(gwr_adaptive, 
          "data/model/gwr_adaptive.rds")
```





### Retrieve gwr output object
The code chunk below will be used to retrieve the save gwr model object.
```{r}
#| include: false
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
```

The code below can be used to display the model output.

```{r}
gwr_adaptive
```
### Computing adaptive bandwidth for the test data
```{r}
#| eval: false
gwr_bw_test_adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
    WITHIN_1KM_PRISCH,
  data=test_data,
  approach="CV",
  kernel="gaussian",
  adaptive=TRUE,
  longlat=FALSE)

write_rds(gwr_bw_test_adaptive, 
          "data/model/gwr_bw_test_adaptive.rds")
```
```{r}
#| include: false
gwr_bw_test_adaptive <- read_rds("data/model/gwr_bw_test_adaptive.rds")
```

```{r}
gwr_bw_test_adaptive
```


### Computing predicted values of the test data
```{r}
#| eval: false
gwr_pred <- gwr.predict(
  formula = resale_price ~ floor_area_sqm + 
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER + 
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
    WITHIN_1KM_PRISCH, 
  data=train_data, 
  predictdata = test_data, 
  bw=80, 
  kernel = 'gaussian', 
  adaptive=TRUE, 
  longlat = FALSE)





write_rds(gwr_pred, 
          "data/model/gwr_pred.rds")

```

```
`