---
title: "Hands-on Exercise 3"
author: "Gautamgovan"
date: "Sep 4, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    number-sections: true
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true 
---
## Overview
A spatio-temporal point process (also called space-time or spatial-temporal point process) is a random collection of points, where each point represents the time and location of an event. Examples of events include incidence of disease, sightings or births of a species, or the occurrences of fires, earthquakes, lightning strikes, tsunamis, or volcanic eruptions.

The analysis of spatio-temporal point patterns is becoming increasingly necessary, given the rapid emergence of geographically and temporally indexed data in a wide range of fields. Several spatio-temporal point patterns analysis methods have been introduced and implemented in R in the last ten years. This chapter shows how various R packages can be combined to run a set of spatio-temporal point pattern analyses in a guided and intuitive way. A real world forest fire events in Kepulauan Bangka Belitung, Indonesia from 1st January 2023 to 31st December 2023 is used to illustrate the methods, procedures and interpretations.

The specific questions we would like to answer are:

- Are the locations of forest fire in Kepulauan Bangka Belitung spatial and spatio-temporally independent?
- If the answer is **NO**, where and when do the observed forest fire locations tend to cluster?

## Data 

For the purpose of this exercise, two data sets will be used, they are:

- **forestfires**, a CSV file providing locations of forest fires detected from the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor data.  
  The data are downloaded from [Fire Information for Resource Management System](https://firms.modaps.eosdis.nasa.gov/).  
  

- **Kepulauan_Bangka_Belitung**, an ESRI shapefile showing the sub-district (i.e. *kelurahan*) boundary of Kepulauan Bangka Belitung.  
  The data set was downloaded from [Indonesia Geospatial](https://tanahair.indonesia.go.id/portal-web) portal.  
  The original data covers the whole Indonesia. For the purpose of this exercise, only sub-districts within Kepulauan Bangka Belitung are extracted.
  
## Installing and Loading the R packages


For the purpose of this study, six R packages will be used. They are:

- **[sf](https://r-spatial.github.io/sf/)** – provides functions for importing, processing, and wrangling geospatial data.
- **[raster](https://rspatial.org/raster/)** – for handling raster data in R.
- **[spatstat](https://spatstat.org/)** – for performing Spatial Point Patterns Analysis such as `Kcross`, `Lcross`, etc.
- **[sparr](https://cran.r-project.org/package=sparr)** – provides functions to estimate fixed and adaptive kernel-smoothed spatial relative risk surfaces via the density-ratio method and perform subsequent inference. Fixed-bandwidth spatiotemporal density and relative risk estimation is also supported.
- **[tmap](https://cran.r-project.org/package=tmap)** – provides functions to produce cartographic quality thematic maps.
- **[tidyverse](https://www.tidyverse.org/)** – a family of R packages that provide functions to perform common data science tasks including, but not limited to, data import, data transformation, data wrangling, and data visualisation.

```{r}
pacman::p_load(sf, terra, spatstat, sparr, tmap, tidyverse)
```
## Importing and Preparing Study Area

### Importing study area

```{r}

kbb_sf <- st_read(
  dsn   = "C:/govanzz/ISSS626_GEO_AUG_2025/Hands-on_Ex/Hands-on_Ex03/Data/geospatial",
  layer = "Kepulauan_Bangka_Belitung"
) %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)   # WGS 84 / UTM zone 48S

```
### Converting OWIN
Next, as.owin() is used to convert kbb into an owin object.

```{r}
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```
Next, class() is used to confirm if the output is indeed an owin object.
```{r}
class(kbb_owin)
```
## Importing and Preparing Forest Fire data

Next, we will import the forest fire data set (i.e. forestfires.csv) into R environment.

```{r}

fire_sf <- read_csv("C:/govanzz/ISSS626_GEO_AUG_2025/Hands-on_Ex/Hands-on_Ex03/Data/aspatial/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326) %>%
  st_transform(crs = 32748)
```
```{r}
fire_sf <- fire_sf %>% 
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, 
                           label = TRUE, 
                           abbr = FALSE))
```

## Visualising the Fire Points





















