---
title: "Hands on Exercise 05"
author: "Gautamgovan"
date: "Sep 25, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 4 
    number-sections: true
    
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true 
---
## Global Measures of Spatial Autocorrelation

### Overview

In this hands-on exercise, we will learn how to compute **Global Measures of Spatial Autocorrelation (GMSA)** using the **spdep** package.  
By the end of this exercise, you will be able to:

- Import geospatial data using appropriate functions from the **sf** package.
- Import CSV files using appropriate functions from the **readr** package.
- Perform relational joins using appropriate join functions from the **dplyr** package.
- Compute Global Spatial Autocorrelation (GSA) statistics using functions from the **spdep** package:
  - Plot a Moran scatterplot.
  - Compute and plot a spatial correlogram using functions from the **spdep** package.
- Provide a statistically correct interpretation of GSA statistics.

### Getting Started

#### The analytical question
In spatial policy, one of the main development objective of the local government and planners is to ensure equal distribution of development in the province. Our task in this study, hence, is to apply appropriate spatial statistical methods to discover if development are even distributed geographically. If the answer is No. Then, our next question will be “is there sign of spatial clustering?”. And, if the answer for this question is yes, then our next question will be “where are these clusters?”

In this case study, we are interested to examine the spatial pattern of a selected development indicator (i.e. GDP per capita) of Hunan Provice, People Republic of China.

#### The Study Area
Two data sets will be used in this hands-on exercise, they are:

- Hunan province administrative boundary layer at county level. This is a geospatial data set in ESRI shapefile format.
- Hunan_2012.csv: This csv file contains selected Hunan’s local development indicators in 2012.

#### Setting the Analytical Tools

Before we get started, we need to ensure that the **spdep**, **sf**, **tmap**, and **tidyverse** packages of R are installed in your environment.

- **sf**: for importing and handling geospatial data in R.  
- **tidyverse**: for wrangling attribute data in R.  
- **spdep**: for computing spatial weights and performing global and local spatial autocorrelation analysis.  
- **tmap**: for preparing cartographic-quality choropleth maps.

The following code chunk will:

- Create a package list containing the required R packages.
- Check whether the packages in the list are installed:
  - If not installed, RStudio will install the missing packages automatically.
- Load the packages into the R environment.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

### Getting the Data Into R Environment
In this section, you will learn how to bring a geospatial data and its associated attribute table into R environment. The geospatial data is in ESRI shapefile format and the attribute table is in csv fomat.

####  Import shapefile into r environment

The code chunk below uses st_read() of sf package to import Hunan shapefile into R. The imported shapefile will be simple features Object of sf.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```
#### Import csv file into r environment

Next, we will import Hunan_2012.csv into R by using read_csv() of readr package. The output is R data frame class.
```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

#### Performing relational join
The code chunk below will be used to update the attribute table of hunan’s SpatialPolygonsDataFrame with the attribute fields of hunan2012 dataframe. This is performed by using left_join() of dplyr package.

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```
#### Visualising Regional Development Indicator
Now, we are going to prepare a basemap and a choropleth map showing the distribution of GDPPC 2012 by using qtm() of tmap package.

```{r}
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```
### Global Measures of Spatial Autocorrelation
In this section, we will learn how to compute global spatial autocorrelation statistics and to perform spatial complete randomness test for global spatial autocorrelation.

#### Computing Contiguity Spatial Weights
Before we can compute the global spatial autocorrelation statistics, we need to construct a spatial weights of the study area. The spatial weights is used to define the neighbourhood relationships between the geographical units (i.e. county) in the study area.

In the code chunk below, poly2nb() of spdep package is used to compute contiguity weight matrices for the study area. This function builds a neighbours list based on regions with contiguous boundaries. If you look at the documentation you will see that you can pass a “queen” argument that takes TRUE or FALSE as options. If you do not specify this argument the default is set to TRUE, that is, if you don’t specify queen = FALSE this function will return a list of first order neighbours using the Queen criteria.

More specifically, the code chunk below is used to compute Queen contiguity weight matrix.

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```
The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one neighbours.

#### Row-standardised weights matrix

Next, we need to assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned equal weight (style=“W”). This is accomplished by assigning the fraction 1/(#ofneighbors) to each neighboring county then summing the weighted income values. While this is the most intuitive way to summaries the neighbors’ values it has one drawback in that polygons along the edges of the study area will base their lagged values on fewer polygons thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data. For this example, we’ll stick with the style=“W” option for simplicity’s sake but note that other more robust options are available, notably style=“B”.

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

### Global Measures of Spatial Autocorrelation: Moran’s I

In this section, we will learn how to perform **Moran’s I** statistics testing by using [`moran.test()`](https://r-spatial.github.io/spdep/reference/moran.test.html) of **spdep**.

#### Maron’s I test

The code chunk below performs Moran’s I statistical testing using [`moran.test()`](https://r-spatial.github.io/spdep/reference/moran.test.html) of spdep.

```{r}
moran.test(hunan$GDPPC, 
           listw=rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```
The Moran’s I test shows I = 0.3007 with z = 4.735 and p = 1.095e-06 (alternative = “greater”). 
We therefore reject the null hypothesis of no spatial autocorrelation and conclude there is significant positive spatial autocorrelation in GDP per capita across Hunan counties—neighboring counties tend to have similar GDPPC values (high-with-high, low-with-low). 
The magnitude (~0.30) suggests moderate clustering rather than randomness.

#### Computing Monte Carlo Moran’s I

The code chunk below performs a permutation test for **Moran’s I** statistic by using [`moran.mc()`](https://r-spatial.github.io/spdep/reference/moran.mc.html) of **spdep**. A total of **1000 simulations** will be performed.
```{r}
set.seed(1234)
bperm= moran.mc(hunan$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm
```
The permutation test shows a Moran’s I = 0.30075 with a p-value = 0.001, which is far below the typical significance level (e.g., 0.05). 

- Therefore, we reject the null hypothesis of spatial randomness.

- This indicates that GDP per capita values across Hunan counties exhibit significant      positive spatial autocorrelation, meaning areas with high (or low) GDPPC tend to         cluster near each other rather than being randomly distributed.

####  Visualising Monte Carlo Moran’s I

It is always good practice to examine the simulated **Moran’s I** test statistics in greater detail. This can be done by plotting the distribution as a histogram using the code chunk below.

In the code below, base R’s [`hist()`](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/hist.html) and [`abline()`](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/abline.html) are used.

```{r}
mean(bperm$res[1:999])
```
```{r}
var(bperm$res[1:999])
```
```{r}
summary(bperm$res[1:999])
```
```{r}
hist(bperm$res, 
     freq=TRUE, 
     breaks=20, 
     xlab="Simulated Moran's I")
abline(v=0, 
       col="red") 
```

**Statistical observation**

- The permutation distribution of Moran’s I is centered near **0** (mean ≈ **–0.015**, var ≈ **0.00437**), as expected under spatial randomness.
- The **observed Moran’s I = 0.30075** lies far in the **right tail** of this distribution.
- The Monte-Carlo **p-value = 0.001**, so we **reject H₀ (no spatial autocorrelation)** and conclude **significant positive spatial autocorrelation** in GDPPC.


**ggplot2 version of the histogram**

```{r}
library(ggplot2)

# sims: the permutation distribution; obs: observed Moran's I
sims <- bperm$res[1:999]                 # 999 simulated values
obs  <- as.numeric(bperm$statistic)      # observed Moran's I
p_emp <- mean(sims >= obs)               # empirical one-sided p-value

df <- data.frame(sim = sims)

ggplot(df, aes(x = sim)) +
  geom_histogram(bins = 30, fill = "grey40", color = "black") +
  geom_vline(xintercept = 0, color = "red", linewidth = 1) +
  geom_vline(xintercept = obs, linetype = "dashed", color = "blue", linewidth = 1) +
  expand_limits(x = obs) +
  labs(
    title = "Permutation Distribution of Moran's I",
    x = "Simulated Moran's I",
    y = "Count",
    subtitle = paste0("Observed I = ", round(obs, 5),
                      " | Empirical p = ", signif(p_emp, 3))
  ) +
  theme_minimal()

```
### Global Measures of Spatial Autocorrelation: Geary’s C

In this section, we will learn how to perform Geary’s C statistics testing by using appropriate functions of spdep package.

#### Geary's C test
The code chunk below performs Geary’s C test for spatial autocorrelation by using geary.test() of spdep.

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```
**Statistical conclusion**

- The Geary’s C test produced a statistic of **0.6907**, which is **significantly lower than 1** (the expected value under spatial randomness).  
- The associated **p-value = 0.0001526** is much less than 0.05, so we **reject the null hypothesis** of no spatial autocorrelation.  
- A Geary’s C value below 1 indicates **positive spatial autocorrelation**, meaning that counties with similar GDP per capita values are spatially clustered together.  
- This result supports the conclusion that **GDPPC values are not randomly distributed**, but instead show **spatial dependence and similarity among neighboring regions**.

#### Computing Monte Carlo Geary’s C
The code chunk below performs permutation test for Geary’s C statistic by using geary.mc() of spdep.
```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
```

**Statistical conclusion**

- The Monte Carlo permutation test for Geary’s C produced a statistic of **0.69072**, which is significantly lower than the expected value of **1** under spatial randomness.  
- The **p-value = 0.001** indicates strong evidence against the null hypothesis of no spatial autocorrelation.  
- Since the observed Geary’s C is less than 1, we conclude that there is **significant positive spatial autocorrelation** in GDP per capita.  
- This means that regions with similar GDPPC values tend to be **spatially clustered**, rather than randomly distributed across Hunan province.

#### Visualising the Monte Carlo Geary’s C

Next, we will plot a histogram to reveal the distribution of the simulated values by using the code chunk below.

```{r}
mean(bperm$res[1:999])
```
```{r}
var(bperm$res[1:999])
```
```{r}
summary(bperm$res[1:999])
```
```{r}
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red") 
```

**Statistical observation**

- The permutation distribution of Geary’s C is centered near **1.0044**, which is close to the expected value of **1** under spatial randomness.  
- The observed Geary’s C statistic (**0.69072**) is located far to the **left tail** of the simulated distribution, indicating it is significantly lower than expected by chance.  
- Since the observed statistic is much less than 1, we **reject the null hypothesis** of no spatial autocorrelation.  
- This provides strong evidence of **positive spatial autocorrelation**, meaning that regions with similar GDP per capita values are **spatially clustered** rather than randomly distributed.

### Spatial Correlogram
Spatial correlograms are great to examine patterns of spatial autocorrelation indata or model residuals. They show how correlated are pairs of spatial observations when you increase the distance (lag) between them - they are plots of some index of autocorrelation (Moran’s I or Geary’s c) against distance.Although correlograms are not as fundamental as variograms (a keystone concept of geostatistics), they are very useful as an exploratory and descriptive tool. For this purpose they actually provide richer information than variograms.

#### Compute Moran’s I correlogram

In the code chunk below, sp.correlogram() of spdep package is used to compute a 6-lag spatial correlogram of GDPPC. The global spatial autocorrelation used in Moran’s I. The plot() of base Graph is then used to plot the output.

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)
```
By plotting the output might not allow us to provide complete interpretation. This is because not all autocorrelation values are statistically significant. Hence, it is important for us to examine the full analysis report by printing out the analysis results as in the code chunk below.

```{r}
print(MI_corr)
```
**Statistical observation**

- The Moran’s I correlogram shows how spatial autocorrelation of GDP per capita changes with increasing spatial lags.  
- For the first and second lags, Moran’s I values (**0.3008** and **0.2060**) are **positive and statistically significant (p < 0.001)**, indicating **strong positive spatial autocorrelation** — nearby regions tend to have similar GDPPC values.  
- At the third lag (**0.0668**), Moran’s I remains positive but the strength of autocorrelation decreases and is only **marginally significant**.  
- Beyond the fourth lag, Moran’s I values become **insignificant or negative**, with significant negative values at the fifth (**–0.1530**) and sixth (**–0.1187**) lags, suggesting **spatial dispersion or dissimilarity** at larger distances.  
- Overall, this pattern indicates that **positive spatial clustering is strongest at short distances**, but **weakens and turns negative as distance increases**, revealing a spatial structure where similarity diminishes with spatial separation.

#### Compute Geary’s C correlogram and plot
In the code chunk below, sp.correlogram() of spdep package is used to compute a 6-lag spatial correlogram of GDPPC. The global spatial autocorrelation used in Geary’s C. The plot() of base Graph is then used to plot the output.

```{r}

GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)
```
Similar to the previous step, we will print out the analysis report by using the code chunk below.
```{r}
print(GC_corr)
```
**Statistical conclusion**

- The Geary’s C correlogram provides insight into how spatial autocorrelation changes across increasing spatial lags.  
- At lower lags (e.g., lag 1 and 2), Geary’s C values are **significantly less than 1**, indicating **strong positive spatial autocorrelation** — nearby regions tend to have similar GDP per capita values.  
- As the spatial lag increases, the Geary’s C statistic approaches or exceeds **1**, suggesting that spatial autocorrelation weakens and may transition towards randomness or even **negative autocorrelation**.  
- This pattern confirms that spatial dependence is strongest at **short distances** and gradually diminishes as the spatial separation between regions increases.

##  Local Measures of Spatial Autocorrelation

### Overview

Local Measures of Spatial Autocorrelation (LMSA) focus on the relationships between each observation and its surroundings, rather than providing a single summary of these relationships across the map. In this sense, they are not summary statistics but scores that allow us to learn more about the spatial structure in our data. The general intuition behind the metrics however is similar to that of global ones. Some of them are even mathematically connected, where the global version can be decomposed into a collection of local ones. One such example are Local Indicators of Spatial Association (LISA). Beside LISA, Getis-Ord’s Gi-statistics will be introduce as an alternative LMSA statistics that present complementary information or allow us to obtain similar insights for geographically referenced data.

In this hands-on exercise, we will learn how to compute **Local Measures of Spatial Autocorrelation (LMSA)** by using the [`sfdep`](https://r-spatial.github.io/sfdep/) package. By the end of this hands-on exercise, you will be able to:

- Import geospatial data using appropriate function(s) of the **sf** package  
- Import CSV files using appropriate functions of the **readr** package  
- Perform relational joins using appropriate join functions of the **dplyr** package  
- Compute **Local Indicator of Spatial Association (LISA)** statistics for detecting clusters and outliers using appropriate functions of the **sfdep** package  
- Compute **Getis-Ord’s Gi-statistics** for detecting hot spot and/or cold spot areas using appropriate functions of the **sfdep** package  
- Visualise the analysis output using the **tmap** package

### Getting Started 

#### The analytical question

In spatial policy, one of the main development objective of the local govenment and planners is to ensure equal distribution of development in the province. Our task in this study, hence, is to apply appropriate spatial statistical methods to discover if development are even distributed geographically. If the answer is No. Then, our next question will be “is there sign of spatial clustering?”. And, if the answer for this question is yes, then our next question will be “where are these clusters?”

In this case study, we are interested to examine the spatial pattern of a selected development indicator (i.e. GDP per capita) of Hunan Provice, People Republic of China.(https://en.wikipedia.org/wiki/Hunan)

#### Study Area and Data

Two data sets will be used in this hands-on exercise, they are:

- Hunan province administrative boundary layer at county level. This is a geospatial data set in ESRI shapefile format.
- Hunan_2012.csv: This csv file contains selected Hunan’s local development indicators in 2012.

#### Setting the Analytical Tools
Before we get started, we need to ensure that the **spdep**, **sf**, **tmap**, and **tidyverse** packages of R are installed in your environment.

- **sf**: for importing and handling geospatial data in R.  
- **tidyverse**: for wrangling attribute data in R.  
- **spdep**: for computing spatial weights and performing global and local spatial autocorrelation analysis.  
- **tmap**: for preparing cartographic-quality choropleth maps.

The following code chunk will:

- Create a package list containing the required R packages.
- Check whether the packages in the list are installed:
  - If not installed, RStudio will install the missing packages automatically.
- Load the packages into the R environment.

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

#### Import shapefile into r environment
The code chunk below uses st_read() of sf package to import Hunan shapefile into R. The imported shapefile will be simple features Object of sf.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan") %>% 
  st_transform(crs = 32650)
```
#### Import csv file into r environment

Next, we will import Hunan_2012.csv into R by using read_csv() of readr package. The output is R data frame class.

```{r}

hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

#### Performing relational join

The code chunk below will be used to update the attribute table of hunan’s sf data frame with the attribute fields of hunan2012 dataframe. This is performed by using left_join() of dplyr package.

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```
#### Visualising Regional Development Indicator

- In the code chunks below, tmap functions are used:

- to build two choropleth maps by using equal interval (i.e equal) and quantile (i.e. quantile) classification methods, and
to plot both maps next to each other by using tmap_arrange().

```{r}
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile interval classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

- The choropleth maps above show the distribution of GDP per capita across regions using two classification methods (Equal Interval and Quantile Interval). While they reveal **spatial variation** in GDPPC values, they **do not clearly identify outliers or clusters**. The maps show areas with higher or lower values, but they do not provide statistical evidence of spatial clustering or outlier behavior.

- Similarly, the plots **do not directly indicate the presence of hot spots or cold spots**. They highlight regions with relatively higher or lower GDPPC values, but identifying statistically significant hot or cold spots requires spatial analysis methods such as **Local Moran’s I (LISA)** or **Getis-Ord Gi\*** statistics.

### Local Indicators of Spatial Association(LISA)

Local Indicators of Spatial Association or LISA are statistics that evaluate the existence of clusters and/or outliers in the spatial arrangement of a given variable. For instance if we are studying distribution of GDP per capita of Hunan Provice, People Republic of China, local clusters in GDP per capita mean that there are counties that have higher or lower rates than is to be expected by chance alone; that is, the values occurring are above or below those of a random distribution in space.

In this section, we will learn how to apply appropriate Local Indicators for Spatial Association (LISA), especially local Moran’I to detect cluster and/or outlier from GDP per capita 2012 of Hunan Province, PRC.

#### Computing Contiguity Spatial Weights

Before we can compute the local spatial autocorrelation statistics, we need to construct a **spatial weights matrix** for our study area. The spatial weights are used to define the neighbourhood relationships between the geographical units (e.g., counties) in our study area.

In the code chunk below, we use [`st_contiguity()`](https://r-spatial.github.io/sfdep/reference/st_contiguity.html) from the **sfdep** package to compute contiguity weight matrices. This function builds a neighbours list based on regions with contiguous boundaries.  
We can pass a `queen` argument (TRUE or FALSE) to specify the type of contiguity:

- `queen = TRUE` (default): Uses **Queen contiguity** (shared point or edge).  
- `queen = FALSE`: Uses **Rook contiguity** (shared edge only).

Next, we use [`st_weights()`](https://r-spatial.github.io/sfdep/reference/st_weights.html) to calculate polygon spatial weights from the `nb` list.

The `st_weights()` function takes three main arguments:

- **nb**: A neighbour list object created by [`st_neighbors()`](https://r-spatial.github.io/sfdep/reference/st_neighbors.html).  
- **style**: The weighting style (default is `"W"` for row-standardized weights). Other options include:  
  - `"B"`: Basic binary coding  
  - `"C"`: Globally standardised (sums over all links to *n*)  
  - `"U"`: C divided by the number of neighbours (sums over all links to unity)  
  - `"S"`: Variance-stabilising coding scheme (Tiefelsdorf et al., 1999, pp. 167–168)  
- **allow_zero**: If `TRUE`, assigns zero as a lagged value to zones without neighbours.

```{r}
wm_q <- hunan %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(
           nb, style = "W"),
         .before = 1) 
```


To reveal the neighbour list, code chunk below is used.

```{r}
summary(wm_q$nb)
```
The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one neighbour.

#### Row-standardised weights matrix


Next, we need to assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned equal weight (style=“W”). This is accomplished by assigning the fraction 1/(#ofneighbors) to each neighboring county then summing the weighted income values. While this is the most intuitive way to summaries the neighbors’ values it has one drawback in that polygons along the edges of the study area will base their lagged values on fewer polygons thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data. For this example, we’ll stick with the style=“W” option for simplicity’s sake but note that other more robust options are available, notably style=“B”.

```{r}
wm_q <- hunan %>%
  mutate(
    nb = st_contiguity(geometry),
    wt = st_weights(nb, style = "W")
  )
```

#### Computing local Moran’s I
To compute local Moran’s I, the local_moran() function of sfdep will be used. It computes Ii values, given a set of zi values and a listw object providing neighbour weighting information for the polygon associated with the zi values.

The code chunks below are used to compute local Moran’s I of GDPPC2012 at the county level.

```{r}
lisa <- wm_q %>% 
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

local_moran() function returns a matrix of values whose columns are:

- Ii: the local Moran’s I statistics
- E.Ii: the expectation of local moran statistic under the randomisation hypothesis
- Var.Ii: the variance of local moran statistic under the randomisation hypothesis
- Z.Ii:the standard deviate of local moran statistic
- Pr(): the p-value of local moran statistic

```{r}
glimpse(lisa)
```
##### Mapping local Moran’s I values

Using choropleth mapping functions of tmap package, we can plot the local Moran’s I values by using the code chinks below.

```{r}
tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Morans'I",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Morans'I of GDPPC (Queen's method)")
```
#####  Mapping local Moran’s I p-values
The choropleth shows there is evidence for both positive and negative Ii values. However, it is useful to consider the p-values for each of these values, as consider above.

The code chunks below produce a choropleth map of Moran’s I p-values by using functions of tmap package.


```{r}
tm_shape(lisa) +
  tm_polygons(fill = "p_ii", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")
```
##### Mapping both local Moran’s I values and p-values
For effective interpretation, it is better to plot both the local Moran’s I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Moran's I of GDPPC (Queen's method)")

p_ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "p_ii", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_in("left", "bottom")
    )) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")

tmap_arrange(ii.map, p_ii.map, asp=1, ncol=2)
```

#### Preparing and Visualising LISA Map

Visualizing LISA involves creating maps that highlight significant spatial patterns, such as hot spots (high values clustered together) and cold spots (low values clustered together), as well as identifying spatial outliers. Common methods include choropleth maps showing LISA values, Moran scatterplots which depict local spatial autocorrelation for individual locations, and hot spot analysis maps.

The LISA Cluster Map shows the significant locations color coded by type of spatial autocorrelation. The first step before we can generate the LISA cluster map is to plot the Moran scatterplot.

#####  Plotting Moran scatterplot
A Moran Scatterplot is a graphical tool in spatial analysis that visualizes the relationship between a variable’s values and the average values of its neighboring locations (its spatial lag) to identify spatial autocorrelation. By plotting standardized values of the variable on the x-axis and its spatial lag on the y-axis, the plot’s slope represents Moran’s I, revealing patterns of high-high or low-low value clusters and high-low or low-high value outliers.

In the code chunk below, st_lag() of sfdep package is used to calculates the spatial lag of GDPPC given a neighbor (i.e. nb) and weights list (i.e. wt).

```{r}
lisa <- lisa %>%
  mutate(lag_GDPPC = st_lag(
    GDPPC, nb, wt),
    .before = 1) %>%
  unnest(lag_GDPPC)

```

Then, the code chunk below is used to plot the Moran Scatterplot of GDPPC against its spatial lag version as shown in the code cunk below.

```{r}
ggplot(data = lisa, 
       aes(x = GDPPC, 
           y = lag_GDPPC)) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "red") +
  labs(x = "GDPPC",
       y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot") +
  theme_minimal()
```
The Moran Scatterplot above is not very intuitive. To support effective visual communication, the code chunk below is used.

```{r}
ggplot(data = lisa, 
       aes(x = GDPPC, 
           y = lag_GDPPC, 
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "black") +
  geom_hline(yintercept=mean(lisa$lag_GDPPC), lty=2) + 
    geom_vline(xintercept=mean(lisa$GDPPC), lty=2) +
  scale_color_manual(
    values = c("High-High" = "red", 
               "Low-Low" = "blue",
               "Low-High" = "lightblue", 
               "High-Low" = "pink")) +
  labs(x = "GDPPC",
       y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

##### Plotting Moran scatterplot with standardised variable
First we will use scale() to centers and scales the variable. Here centering is done by subtracting the mean (omitting NAs) the corresponding columns, and scaling is done by dividing the (centered) variable by their standard deviations.

```{r}
lisa <- lisa %>%
  mutate(z_GDPPC = scale(GDPPC),
         z_lag_GDPPC = scale(lag_GDPPC),
         .before = 1)
```

The as.vector() added to the end is to make sure that the data type we get out of this is a vector, that map neatly into out dataframe.

Now, we are ready to plot the Moran scatterplot again by using the code chunk below.

```{r}
ggplot(data = lisa, 
       aes(x = z_GDPPC, 
           y = z_lag_GDPPC, 
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "black") +
  geom_hline(yintercept=mean(lisa$z_lag_GDPPC), lty=2) + 
    geom_vline(xintercept=mean(lisa$z_GDPPC), lty=2) +
  scale_color_manual(
    values = c("High-High" = "red", 
               "Low-Low" = "blue",
               "Low-High" = "lightblue", 
               "High-Low" = "pink")) +
  labs(x = "Standardised GDPPC",
       y = "Standardised Spatial Lag of GDPPC",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```
##### Preparing LISA map classes
The code chunks below show the steps to prepare a LISA cluster map.

Next, derives the spatially lagged variable of interest (i.e. GDPPC) and centers the spatially lagged variable around its mean.

This is follow by centering the local Moran’s around the mean.

Next, we will set a statistical significance level for the local Moran.

```{r}
signif <- 0.05
```

##### Plotting and visualising LISA map
By default, the LISA classes are stored in mean, median and pysal fields of lisa sf data frame. To prepare the LISA map, we will derive a new field called LISA-cluster from either one of these three field by using the code chunk below.

```{r}
lisa <- lisa %>%
  mutate(
    LISA_cluster = ifelse(
      p_ii < 0.05,
      as.character(mean), 
      "Insignificant"),
    LISA_cluster = factor(
      LISA_cluster,
      levels = c("Insignificant", "Low-Low", "Low-High", "High-Low", "High-High")
    )
  )
```

Now, we can plot the LISA map by using the code chunk below.

```{r}
lisa_map <- tm_shape(lisa) + 
  tm_polygons(
    fill = "LISA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",      # Insignificant
        "blue",        # Low-Low
        "lightblue",   # Low-High
        "pink",        # High-Low
        "red"          # High-High
      )
    ),
    fill.legend = tm_legend(
      title = "LISA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("Local Moran's I Clusters (p < 0.05)")
lisa_map
```
For effective interpretation, it is better to plot both the local Moran’s I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

We can also include the local Moran’s I map and p-value map as shown below for easy comparison.

```{r}
# Create the two maps separately
map1 <- tm_shape(lisa) +
  tm_polygons(
    fill = "ii",
    fill.scale = tm_scale_intervals(
      style = "pretty",
      n = 5,
      values = "brewer.RdBu"
    ),
    fill.legend = tm_legend(
      title = "Local Moran's I",
      position = tm_pos_out()
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Moran's I of GDPPC (Queen's method)")

map2 <- tm_shape(lisa) +
  tm_polygons(
    fill = "LISA_cluster",
    fill.scale = tm_scale_categorical(
      values = c("grey80", "blue", "lightblue", "pink", "red")
    ),
    fill.legend = tm_legend(
      title = "LISA Cluster",
      position = tm_pos_in("left", "bottom")
    )
  ) +
  tm_borders() +
  tm_title("Local Moran's I Clusters (p < 0.05)")

# Combine them side by side
tmap_arrange(map1, map2, ncol = 2)

```

**Statistical observation**

- The Local Moran’s I map (left) shows the spatial distribution of local autocorrelation in GDP per capita. Darker blue regions indicate areas with **high positive spatial autocorrelation**, where counties with high GDPPC are surrounded by similarly high values.
- The LISA cluster map (right) further classifies these relationships into statistically significant clusters.  
  - **High-High clusters (red)**: Regions with high GDPPC surrounded by high GDPPC neighbours — forming economic “hot spots.”  
  - **Low-Low clusters (blue)**: Regions with low GDPPC surrounded by similarly low GDPPC neighbours — forming “cold spots.”  
  - **Low-High (light blue)** and **High-Low (pink)**: Spatial outliers where a region’s GDPPC contrasts with its neighbours.
- Overall, the maps reveal significant **spatial clustering** of economic performance — with clear hot spots in the eastern part of the study area and cold spots in the central-western region.

### Hot Spots and Cold Spots Analysis (HCSA)
Hot spots and cold spots analysis is a technique used in spatial statistics to identify statistically significant clusters of high values (hot spots) and low values (cold spots) of a geographically referenced attribute (i.e. GDPPC). These analyses help visualize and understand spatial patterns, revealing areas where data points are clustered more densely than would be expected by random chance.

Figure below shows the hot spots and cold spots of GDP per capita of Hunan Provice, People Republic of China.

The term hot spots refer to areas where high values are concentrated and surrounded by other high values. It has been used generically across disciplines to describe a region or value that is higher relative to its surroundings (Lepers et al 2005, Aben et al 2012, Isobe et al 2015). For example, a hot spot analysis might reveal clusters of high GDP per capita in a province, indicating areas with significantly higher GDP per capita than other parts of the province.

On the other hands, the term cold spots refer to areas where low values are concentrated and surrounded by other low values. An example could be clusters of low GDP per capita, indicating areas where GDP per capita is consistently low.

HCSA uses statistical methods, like the Getis-Ord Gi* statistic, to determine if the observed clustering is statistically significant or could have occurred randomly.

The analysis consists of four steps:

- Deriving spatial weight matrix
- Computing Gi* statistics
- Mapping Gi* statistics
- Communicating the analysis results

#### Deriving distance weight matrix
Whist the spatial autocorrelation considered units which shared borders, for Getis-Ord Gi* statistics, we are defining neighbours based on distance.

There are two type of distance-based proximity matrix, they are:

- fixed distance weight matrix,
- adaptive distance weight matrix, and
- inverse distance weights.

##### Computing fixed distance weights
In this section, we will learn how to compute fixed distance weight matrix by using st_dist_band() and st_weights() of sfdep package.

Firstly, we need to determine the upper limit for distance band by using the steps below:

```{r}
ct <- critical_threshold(st_geometry(hunan))
ct
```
The summary report shows that the largest first nearest neighbour distance is 60.80 km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

Using the critical threshold value computed above, code chunk below is used to compute the fixed distance weight.

```{r}
hunan_fdw <- hunan %>% 
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry), 
                 upper = ct)),
    wt = st_weights(nb, 
                      style = "W"),
    .before = 1) 
```
##### Computing adaptive distance weights

One of the characteristics of fixed distance weight matrix is that more densely settled areas (usually the urban areas) tend to have more neighbours and the less densely settled areas (usually the rural counties) tend to have lesser neighbours. Having many neighbours smooth the neighbour relationship across more neighbours.

It is possible to control the numbers of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry as shown in the code chunk below.

In this section, we will learn how to compute fixed distance weight matrix by using st_dist_band() and st_weights() of sfdep package.

```{r}
hunan_adw <- hunan %>%
  mutate(nb = include_self(
    st_knn(
      st_geometry(geometry),
      k = 6)),
    wt = st_weights(
      nb, style = "W"),
    .before = 1)
```
#### Computing local Gi*
Local Gi* (pronounced “G-i-star”) is a Getis-Ord statistic used in spatial analysis to identify statistically significant hot spots and cold spots of high or low attribute values within a dataset. It measures the spatial clustering of values by comparing the total value within a specified neighborhood to the global total, producing a Z-score and p-value for each location to determine the likelihood that the observed pattern is not due to random chance.

In this sub-section, we will gain hands-on experience on performing Gi* analysis by using local_gstar_perm() of sfdep package.

```{r}
HCSA_fdw <- hunan_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      GDPPC, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)
```

#### Preparing and Visualising HCSA Map

##### Mapping Gi* with fix distance weights

In the code chunk below, thematic mapping functions of tmaps are used to prepare a choropleth map showing the distribution of Gi* values.

```{r}
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 6,
                values = "brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "Gi*",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Gi* of GDPPC (Fixed Bandwidth d = 60799.91m)")
```
##### Mapping local Gi* p-values
The choropleth shows there is evidence for both positive and negative Ii values. However, it is useful to consider the p-values for each of these values, as consider above.

The code chunk below plots a choropleth map of p-values of Gi* by using functions of tmap package.

```{r}
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "simulated p-value",
      position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (Fixed distance)")
```
##### Mapping both Gi* values and p-values
For effective interpretation, it is better to plot both the Gi* values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
Gi_star_map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Gi* of GDPPC")

p_values_map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim", 
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_in("left", "bottom")
    )) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (fixed distance)")

tmap_arrange(Gi_star_map, p_values_map, asp=1, ncol=2)
```

##### Plotting and visualising HCSA Map

By default, the HCSA classes are stored in cluster field of HCSA_fdw sf data frame. To prepare the HSCA map, we will derive a new field called HCSA_cluster from cluster field by using the code chunk below.
```{r}

HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim > 0.05 ~ "Insignificant",
    p_sim <= 0.05 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.05 & cluster == "Low"  ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Insignificant", "Hot spot", "Cold spot")
    ),
    .before = 1
  )

```

Now, we can plot the HCSA map by using the code chunk below.
```{r}
HCSA_map <- tm_shape(HCSA_fdw) + 
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",      # Insignificant
        "red",        # Low-Low
        "blue"          # High-High
      )
    ),
    fill.legend = tm_legend(
      title = "HSCA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("HCSA Clusters (p < 0.05)")
HCSA_map
```
For effective interpretation, it is better to plot both the local Gi* values map and its corresponding HCSA map next to each other.

The code chunk below will be used to create such visualisation.
```{r}

tmap_arrange(Gi_star_map, HCSA_map, asp=1, ncol=2)

```
**Statistical observation**

- The Local Gi* maps show the spatial distribution of **hot spots** and **cold spots** of GDP per capita across the study area. Areas shaded in **dark red** indicate statistically significant **hot spots** (high GDPPC values surrounded by other high values), while **blue areas** indicate significant **cold spots** (low GDPPC values surrounded by low values).  
- From the first map, we observe that hot spots are primarily concentrated in the **eastern region**, indicating strong clusters of economic activity. These regions have significantly higher GDPPC compared to their neighbours.  
- The cold spots, on the other hand, are concentrated in the **central and western parts** of the study area, indicating clusters of lower economic performance.  
- The second map showing **p-values** confirms the statistical significance of these clusters — darker shades (p < 0.01) highlight areas with highly significant clustering patterns.  
- Overall, the analysis suggests that GDPPC is **not randomly distributed** but shows a clear spatial pattern with **high-value economic hubs** (hot spots) and **low-value clusters** (cold spots), revealing spatial inequality in economic development across the region.
