---
title: "Take Home Exercise 3"
author: "Gautamgovan"
date: "October 28, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 4
    number-sections: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Predictive Modelling of HDB Resale Prices with Geographically Weighted Methods

### Objective

Housing is an essential component of household wealth worldwide. Purchasing a home has always represented a major investment for most individuals. The price of housing is influenced by a wide range of factors. Some are macro-level, such as the overall state of the economy or the inflation rate, while others are property-specific. These can be further categorized into structural and locational factors.

Structural factors refer to the characteristics of the property itself, such as size, fittings, and tenure. Locational factors, on the other hand, relate to the neighbourhood context of the property, including proximity to childcare centres, public transport, and shopping facilities.

Traditionally, housing resale price models have been estimated using the Ordinary Least Squares (OLS) method. However, this approach does not account for spatial autocorrelation and spatial heterogeneity, which are commonly present in geographic datasets such as housing transactions. When spatial autocorrelation exists, OLS estimation of housing resale price models may produce biased, inconsistent, or inefficient results (Anselin, 1998). To address these limitations, Geographically Weighted Models (GWMs) have been introduced to more accurately calibrate both explanatory and predictive models for housing resale prices.

The objective of this take-home exercise is to develop a spatially informed model to predict HDB resale prices in Singapore, making use of appropriate geospatial analytics methods.

We will develop a model to predict HDB resale prices for the period July to September 2025, using HDB resale transaction data from July 2024 to June 2025.

## Installing and Loading R packages

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, Metrics, tidyverse,httr, tidyverse, 
               jsonlite, progress,ggplot2)
```

## Data Preparation

### HDB Dataset

#### Selecting Date of Interest 
This data set comprises of HDB resale transactions from Jan 2017 onwards. For the purpose of the hands on Exercise, the code chunk below will be used to extract a sub-set of transaction records on 5-Room flat transacted from July 2024 to September 2025.

```{r}
HDBresale <- read_csv(
  "data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "5 ROOM") %>%
  filter(month >= "2024-07" & month < "2025-10")
```

#### Geocoding HDB resale data
In this section, we will geocode HDB resale data by using SLA Onemap Geocode API.

Let us prepare a geocoding function by using the code chunk below. It aims to perform the following tasks:

Combining the block and street name into an address. The addresses will be used to geocode instead of postal code that you learned in In-class Exercise 3b,
Passing the address as the searchVal in our query,
Sending the query to OneMapSG search. Note: Since we don’t need all the address details, we can set getAddrDetails as ‘N’,
Converting response (JSON object) to text,
Saving response in text form as a dataframe, and
Retain the latitude and longitude for our output.

```{r}
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address, 
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  res <- GET(base_url, query = query)
  restext<-content(res, as="text")
  
  output <- fromJSON(restext)  %>% 
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)

  return(output)
}
```
Next, the code chunk below will be used to to execute the geocoding function:

```{r}
#| eval: False
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for (i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```
```{r}
#| eval: false
#| include: false
saveRDS(HDBresale, file = "data/rds/HDBresale_geocoded.rds")
```

```{r}
#| include: false
HDBresale <- readRDS("data/rds/HDBresale_geocoded.rds")
```

#### Converting `storey_range` to a Numerical Value (`storey_mid`)

The `storey_range` field in the HDB resale dataset is provided as a text range such as `"07 TO 09"` or `"13 TO 15"`. Since machine learning models require numerical inputs, we convert this range into a single numeric value that approximates the floor level of the unit.

This is done by:

1. **Removing spaces** from the text  
   - Example: `"07 TO 09"` → `"07TO09"`

2. **Splitting the range** into the lower and upper floor values  
   - `"07TO09"` → `"07"` and `"09"`

3. **Calculating the midpoint** (average) of the two values  
   - \[
     \frac{7 + 9}{2} = 8
     \]

4. **Creating a new variable** named `storey_mid` to store this midpoint value.

This numeric representation allows the floor level information to be used effectively in statistical and machine learning models.

5. Remove the original storey_range column, as it is no longer needed.


```{r}
storey_mid <- function(x) {
  x %>%
    str_replace_all(" ", "") %>%    # remove spaces
    str_split("TO") %>%             # split into lower and upper floors
    sapply(\(v) mean(as.numeric(v)))  # convert to numeric and take midpoint
}

HDBresale <- HDBresale %>%
  mutate(storey_mid = storey_mid(storey_range)) %>%  # create numeric floor feature
  select(-storey_range)                              # remove the original text column

```

#### Converting `remaining_lease` to a Numeric Value (`remaining_lease_yrs`)

The `remaining_lease` field is provided as text (e.g., `"87 years 4 months"`). To use this information in modelling, we convert it into a numeric value representing the total lease remaining in **years**.

The function `convert_remaining_lease()`:

1. Extracts the number of **years**.
2. Extracts the number of **months** (or uses 0 if not specified).
3. Converts months into a fraction of a year (`months / 12`).
4. Adds the two together to get a single numeric value.

We then apply the function to create a new column `remaining_lease_yrs` and drop the original text column.

```{r}
convert_remaining_lease <- function(x) {
  years <- as.numeric(str_extract(x, "\\d+(?=\\syear)"))
  months <- str_extract(x, "(?<=years?\\s)\\d+(?=\\smonth)") %>% as.numeric()
  months[is.na(months)] <- 0
  years + months/12
}

HDBresale <- HDBresale %>%
  mutate(remaining_lease_yrs = convert_remaining_lease(remaining_lease)) %>%
  select(-remaining_lease)
```

### Proximity Analysis
In this section, we will learn how to count the number of geographic entities located within a defined distance from each HDB property.

#### Converting to an sf object
Before performing proximity analysis, it is important to ensure that:

- input data sets must be in sf objects, and
- they must be in similar projected coordinates systems.

In the code chunk below,

- st_as_sf() is used to convert HDBresale tibble data.frame to sf object by using values from LONGITUDE and LATITUDE columns to form the geometry features.
- st_transform() is then used to transform the sf object into svy21 projected coordinates system.
- the output sf object is called HDBresale_sf. It is in sf data.frame format.


```{r}
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326) %>%
  st_transform(crs = 3414)
```
#### PreSchool Location 
Next, we will locate and download Pre-Schools Location data from Singapore’s open data portal. There are both geojson and kml version. In this exercise, the kml version will be used.

Then, code chunk below will be used to import the preschool location data into R environment by using st_read() of readr package.

```{r}
preschool = st_read(
  "data/rawdata/PreSchoolsLocation.kml") %>%
  st_transform(crs= 3414)
```
Lastly, code chunk below will be used to count the number of pre-schools located within 350m of each HDB property.

```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(WITHIN_350M_PRESCHOOL = map_int(within_350m, length))
```

#### Proximity to CBD 
The code chunk below  computes the Euclidean (straight-line) distance between each flat and the Downtown Core (CBD) centroid using Singapore’s projected coordinate system (EPSG:3414, SVY21). The resulting columns proximity_to_cbd_m and proximity_to_cbd_km represent the flat’s distance to the CBD in meters and kilometers respectively

```{r}
# --- Define Downtown Core (CBD) coordinates (WGS84) ---
cbd_lon <- 103.85
cbd_lat <- 1.28967

# --- Create CBD point and transform to SVY21 (3414) ---
cbd_wgs84 <- st_sfc(st_point(c(cbd_lon, cbd_lat)), crs = 4326)
cbd_3414  <- st_transform(cbd_wgs84, 3414)

# --- Compute straight-line distance to CBD (meters / km) ---
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    proximity_to_cbd_m  = as.numeric(st_distance(geometry, cbd_3414)[, 1]),
    proximity_to_cbd_km = proximity_to_cbd_m / 1000
  )
```

```{r}
#| include: false
HDBresale_sf <- HDBresale_sf %>%
  select(-proximity_to_cbd_m)
```

#### Bus Stops within 350 metres

Read Bustops shp file and check CRS 

```{r}
busstops <- st_read("data/rawdata/BusStop.shp", quiet = TRUE)%>%
  st_transform(crs= 3414)
st_crs(busstops)  # check the CRS
```
```{r}
# Count number of bus stops within 350 m for each HDB point
within_350m <- st_is_within_distance(HDBresale_sf, busstops, dist = 350)

# Add count column to main dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(WITHIN_350M_BUSSTOP = purrr::map_int(within_350m, length))
```

#### Elder care within 350 metres

Read Eldercare shapefile and check CRS

```{r}
eldercare <- st_read("data/rawdata/ELDERCARE.shp", quiet = TRUE)%>%
  st_transform(crs= 3414)
st_crs(eldercare)  # check the CRS
```
```{r}
within_350m <- st_is_within_distance(HDBresale_sf, eldercare, dist = 350)

#Add count column to main dataset

HDBresale_sf <- HDBresale_sf %>%
mutate(WITHIN_350M_ELDERCARE = purrr::map_int(within_350m, length))
```

#### Proximity to Good Primary Schools

For this part we have collected top 20 Primary Schools from Singapore in a CSV file along with its postal code , we will use One map API to geo-encode using postal code 

The below code chunk gets the latitude and longitude of the school using postal codes 

```{r}
# --- 1) Read the CSV (school_name, postal_code) ---
good_ps <- read_csv("data/rawdata/good_primary_schools.csv",
                    show_col_types = FALSE) %>%
  mutate(postal_code = str_pad(as.character(postal_code), 6, pad = "0"))
```

```{r}
#| eval: false

# --- 2) OneMap postal-code geocoder (returns SVY21 X/Y) ---
onemap_lookup <- function(pc){
  resp <- RETRY("GET", "https://onemap.gov.sg/api/common/elastic/search",
                query = list(searchVal = pc,
                             returnGeom = "Y",
                             getAddrDetails = "Y",
                             pageNum = "1"),
                quiet = TRUE)
  js <- content(resp, as = "parsed")
  if (!is.null(js$found) && js$found > 0) {
    r <- js$results[[1]]
    tibble(POSTAL = r$POSTAL,
           X = as.numeric(r$X),   # SVY21 easting
           Y = as.numeric(r$Y))   # SVY21 northing
  } else {
    tibble(POSTAL = pc, X = NA_real_, Y = NA_real_)
  }
}

# --- 3) Geocode unique postcodes 
pcs <- unique(good_ps$postal_code)
geo_tbl <- map_dfr(pcs, ~ tryCatch(onemap_lookup(.x),
                                   error = function(e) tibble(POSTAL=.x, X=NA_real_, Y=NA_real_)))

write_rds(geo_tbl, "data/rds/good_ps_postal_geocode.rds")

```

```{r}
geo_tbl <- read_rds("data/rds/good_ps_postal_geocode.rds")
# --- 4) Join back & keep geocoded rows; make sf in SVY21 (EPSG:3414) ---
good_ps_sf <- good_ps %>%
  left_join(geo_tbl, by = c("postal_code" = "POSTAL")) %>%
  filter(!is.na(X), !is.na(Y)) %>%
  st_as_sf(coords = c("X","Y"), crs = 3414)  # OneMap X/Y are SVY21
```

Now the below code chunk is used to derive proximity columns to good primary schools 

```{r}
# --- 5) Proximity features  ---
# Count of good schools within 1 km
idx_1km <- st_is_within_distance(HDBresale_sf, good_ps_sf, dist = 1000)
cnt_1km <- lengths(idx_1km)

# Distance to nearest good school (km)
nearest_idx <- st_nearest_feature(HDBresale_sf, good_ps_sf)
nearest_dist_km <- as.numeric(st_distance(HDBresale_sf, good_ps_sf[nearest_idx, ], by_element = TRUE))/1000

# Add to main dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    WITHIN_1KM_GOOD_PS = cnt_1km,
    dist_good_ps_km    = nearest_dist_km
  )
```

#### Proximity to MRT/LRT stations

This code safely loads and processes MRT/LRT station geometries to calculate each HDB flat’s distance to its nearest station.
It first reads the shapefile with error handling and removes invalid geometries.
For any problematic shapes, it extracts a representative point using a safe function (safe_point).
Next, it groups polygons by station name, merges them, and ensures each has one valid point in SVY21 coordinates.
Finally, it computes straight-line distances (in kilometers) from each HDB flat to its nearest station and adds this as dist_mrt_km to the dataset.



```{r}

# 1) Try reading; if a few polygons fail, catch and skip
mrt_raw <- tryCatch(
  st_read("data/rawdata/RapidTransitSystemStation.shp", quiet = TRUE),
  error = function(e) {
    message("Some polygons failed to load; attempting partial recovery")
    st_read("data/rawdata/RapidTransitSystemStation.shp", quiet = TRUE, stringsAsFactors = FALSE)
  }
)

# 2) Drop any completely invalid geometries
mrt_valid <- mrt_raw[st_is_valid(mrt_raw), ]

# 3) For any remaining invalid shapes, use try() on point extraction
safe_point <- function(geom) {
  pt <- try(st_point_on_surface(geom), silent = TRUE)
  if (inherits(pt, "try-error")) NA else pt
}

# 4) Create one point per station safely
mrt_pts <- mrt_valid %>%
  group_by(STN_NAM_DE) %>%
  summarise(geometry = {
    # combine all parts and get a safe point
    geom <- st_combine(geometry)
    safe_point(geom)
  }, .groups = "drop") %>%
  filter(!st_is_empty(geometry))

st_crs(mrt_pts) <- 3414  # ensure CRS SVY21

# 5) Compute distance to nearest MRT/LRT (km)
nn_idx <- st_nearest_feature(HDBresale_sf, mrt_pts)
dist_mrt_km <- as.numeric(st_distance(HDBresale_sf, mrt_pts[nn_idx, ], by_element = TRUE)) / 1000

# 6) Add to HDB dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(dist_mrt_km = dist_mrt_km)

```

#### Proximity to Parks
Read ParkFacilities (GeoJSON) and ensure CRS
```{r}
parks <- st_read("data/rawdata/ParkFacilities.geojson", quiet = TRUE) %>%
  st_transform(3414)               # SVY21 metres
```

Compute distance to the nearest park (km) and add to dataset
```{r}
# Distance to nearest park facility (km)
nn_idx <- st_nearest_feature(HDBresale_sf, parks)
dist_park_km <- as.numeric(
  st_distance(HDBresale_sf, parks[nn_idx, ], by_element = TRUE)
) / 1000

# Add to main dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(dist_park_km = dist_park_km)
```

#### Proximity to Malls
Read the malls CSV (mall_name, postal_code)
```{r}
malls <- readr::read_csv("data/rawdata/singapore_malls_postal_codes.csv",
                         show_col_types = FALSE) %>%
  dplyr::mutate(postal_code = stringr::str_pad(as.character(postal_code), 6, pad = "0"))
```

Geocode mall postal codes with OneMap (SVY21 X/Y)

```{r}
#| eval: false
#| 
onemap_lookup <- function(pc){
  resp <- httr::RETRY("GET", "https://onemap.gov.sg/api/common/elastic/search",
                      query = list(searchVal = pc,
                                   returnGeom = "Y",
                                   getAddrDetails = "Y",
                                   pageNum = "1"),
                      quiet = TRUE)
  js <- httr::content(resp, as = "parsed")
  if (!is.null(js$found) && js$found > 0) {
    r <- js$results[[1]]
    tibble::tibble(POSTAL = r$POSTAL,
                   X = as.numeric(r$X),   # SVY21 easting
                   Y = as.numeric(r$Y))   # SVY21 northing
  } else {
    tibble::tibble(POSTAL = pc, X = NA_real_, Y = NA_real_)
  }
}

pcs <- unique(malls$postal_code)
mall_geo_tbl <- purrr::map_dfr(
  pcs,
  ~ tryCatch(onemap_lookup(.x),
             error = function(e) tibble::tibble(POSTAL=.x, X=NA_real_, Y=NA_real_))
)

readr::write_rds(mall_geo_tbl, "data/rds/malls_postal_geocode.rds")
```

Join geocodes and create an sf layer (SVY21 / EPSG:3414)

```{r}
mall_geo_tbl <- readr::read_rds("data/rds/malls_postal_geocode.rds")

malls_sf <- malls %>%
  dplyr::left_join(mall_geo_tbl, by = c("postal_code" = "POSTAL")) %>%
  dplyr::filter(!is.na(X), !is.na(Y)) %>%
  sf::st_as_sf(coords = c("X","Y"), crs = 3414)
```

Derive proximity features to malls
```{r}
# Distance to nearest mall (km)
mall_nn_idx <- sf::st_nearest_feature(HDBresale_sf, malls_sf)
dist_mall_km <- as.numeric(sf::st_distance(HDBresale_sf, malls_sf[mall_nn_idx, ], by_element = TRUE)) / 1000

# Count of malls within 1 km (adjust if you prefer another radius)
malls_within_1km <- sf::st_is_within_distance(HDBresale_sf, malls_sf, dist = 1000)
cnt_malls_1km <- lengths(malls_within_1km)

# Add to main dataset
HDBresale_sf <- HDBresale_sf %>%
  dplyr::mutate(
    dist_mall_km       = dist_mall_km,
    WITHIN_1KM_MALLS   = cnt_malls_1km
  )
```

#### Proximity to CHAS clinics
Read Chas data and ensure CRS
```{r}
chas <- st_read("data/rawdata/CHASClinics.geojson", quiet = TRUE) %>%
  st_transform(3414)
```

```{r}
# Distance to nearest CHAS clinic (km)
chas_nn_idx <- st_nearest_feature(HDBresale_sf, chas)
dist_chas_km <- as.numeric(st_distance(HDBresale_sf, chas[chas_nn_idx, ], by_element = TRUE)) / 1000

# Count of CHAS clinics within 350 m
chas_within_350m <- st_is_within_distance(HDBresale_sf, chas, dist = 350)
cnt_chas_350m <- lengths(chas_within_350m)

# Add to main dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    dist_chas_km    = dist_chas_km,
    WITHIN_350M_CHAS = cnt_chas_350m
  )

```

#### Proximity to Hawker Centres


Read Hawker Centres GeoJSON and check CRS 
```{r}
hawker <- st_read("data/rawdata/HawkerCentresGEOJSON.geojson", quiet = TRUE)%>%
  st_transform(3414)

st_crs(hawker)
```
Compute distance to nearest hawker centre (in km) ---
```{r}
hawker_nn_idx <- st_nearest_feature(HDBresale_sf, hawker)
dist_hawker_km <- as.numeric(st_distance(HDBresale_sf, hawker[hawker_nn_idx, ], by_element = TRUE)) / 1000

#  Add distance feature to HDB resale dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(dist_hawker_km = dist_hawker_km)
```

#### Childcare Centres within 350 metres
Read ChildCare Services (GeoJSON) and ensure SVY21:

```{r}
childcare <- st_read("data/rawdata/ChildCareServices.geojson", quiet = TRUE)%>%st_transform(3414)

st_crs(hawker)

```

Compute count within 350 m and add to main data
```{r}
within_350m_cc <- st_is_within_distance(HDBresale_sf, childcare, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(WITHIN_350M_CHILDCARE = lengths(within_350m_cc))
```


```{r}
#| eval: false
#| include: false
colnames(HDBresale_sf)
```
### Calculate Correlation Matrix 
Before loading the predictors into a predictive model, it is always a good practice to use correlation matrix to examine if there is sign of multicollinearity. 
```{r}
# 1) Drop geometry
HDBresale_nogeo <- HDBresale_sf %>%
  st_drop_geometry()

# 2) Select only numeric columns for correlation
numeric_cols <- HDBresale_nogeo %>%
  select(where(is.numeric))

# 3) Compute correlation matrix
corr_matrix <- cor(numeric_cols, use = "complete.obs")

# 4) Plot correlation matrix
corrplot::corrplot(corr_matrix,
                   diag = FALSE,
                   order = "AOE",
                   tl.pos = "td",
                   tl.cex = 0.6,
                   method = "number",
                   type = "upper")
```
Drop Highly correlated Values 
```{r}
HDBresale_sf <- HDBresale_sf %>%
  select(-WITHIN_350M_PRESCHOOL,
         -lease_commence_date,
         -WITHIN_1KM_MALLS,
         -WITHIN_1KM_GOOD_PS)
```

### Train–Test Split for HDB Resale Modelling

This code  chunk separates the dataset into **training** and **testing** periods and removes geometry for modelling.

1. Training Set (Jul 2024 – Jun 2025)
Filters `HDBresale_sf` to include resale transactions from July 2024 to June 2025, storing it as `HDB_train_sf`.

2. Test Set (Jul 2025 – Sep 2025)
Filters the same dataset for resale transactions from July 2025 to September 2025, creating `HDB_test_sf`.

3. Drop Geometry for Modelling
Converts both `sf` objects into regular data frames (`hdb_train`, `hdb_test`) by removing spatial geometry columns using `st_drop_geometry()`, preparing them for statistical or machine-learning models.

```{r}

# Training: Jul 2024 – Jun 2025
HDB_train_sf <- HDBresale_sf %>%
  dplyr::filter(month >= "2024-07" & month <= "2025-06")

# Test (prediction period): Jul 2025 – Sep 2025
HDB_test_sf <- HDBresale_sf %>%
  dplyr::filter(month >= "2025-07" & month <= "2025-09")

# For modelling (no geometry)
hdb_train <- HDB_train_sf %>% sf::st_drop_geometry()
hdb_test  <- HDB_test_sf  %>% sf::st_drop_geometry()
```

```{r}

ggplot(hdb_train, aes(x = resale_price)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of HDB Resale Prices (Raw)",
    x = "Resale Price (SGD)",
    y = "Count"
  ) +
  theme_minimal()
```
## Building a non-spatial multiple linear regression
This code chunk fits a linear regression model (`lm_ols`) using the `lm()` function, where the log-transformed HDB resale price is the dependent variable. The model includes structural attributes (e.g., floor area, storey level, remaining lease, flat model) and locational factors such as proximity to CBD, MRT stations, parks, malls, and amenities like bus stops, childcare, eldercare, and CHAS clinics. The independent variables are selected based on their potential influence on resale price. The model is trained on the `hdb_train` dataset.
```{r}
lm_ols <- lm(
  log(resale_price) ~ 
    floor_area_sqm +
    storey_mid +
    remaining_lease_yrs +
    flat_model +
    proximity_to_cbd_km +
    WITHIN_350M_BUSSTOP +
    WITHIN_350M_ELDERCARE +
    dist_good_ps_km +
    dist_mrt_km +
    dist_park_km +
    dist_mall_km +
    dist_hawker_km +
    WITHIN_350M_CHILDCARE +
    WITHIN_350M_CHAS,
  data = hdb_train
)
```

### OLS Residual Diagnostics & Spatial Dependence Check
#### Summary of the model 
```{r}
summary(lm_ols)
```
The OLS model explains resale prices well, with strong effects from floor area, storey height, lease, and proximity to key amenities. Significant spatial patterns remain, motivating geographically weighted modelling.

```{r}
# Add residuals and run standard OLS diagnostics
hdb_train$residuals <- residuals(lm_ols)

par(mfrow = c(2, 2))
plot(lm_ols)   # includes Residuals vs Fitted, Q-Q plot, etc.

# Check multicollinearity
car::vif(lm_ols)
```
#### Spatial autocorrelation of residuals (Moran’s I)


```{r}

# Build neighbour structure and spatial weights
coords <- st_coordinates(HDB_train_sf)
nb     <- knn2nb(knearneigh(coords, k = 8))
lw     <- nb2listw(nb, style = "W")

# Test for spatial autocorrelation in OLS residuals
moran.test(hdb_train$residuals, lw)
```
OLS residuals exhibit strong and significant spatial autocorrelation (Moran’s I = 0.668, p < 0.001), indicating substantial spatial heterogeneity and justifying the use of geographically weighted models.

### Predict on the test set (Jul–Sep 2025)

```{r}
hdb_test$pred_log <- predict(lm_ols, newdata = hdb_test)
hdb_test$pred_price <- exp(hdb_test$pred_log)
```

### Evaluate predictive performance
```{r}
library(Metrics)

rmse_ols <- rmse(hdb_test$resale_price, hdb_test$pred_price)
mae_ols  <- mae(hdb_test$resale_price, hdb_test$pred_price)
r2_ols   <- 1 - sum((hdb_test$resale_price - hdb_test$pred_price)^2) /
                 sum((hdb_test$resale_price - mean(hdb_test$resale_price))^2)

rmse_ols
mae_ols
r2_ols
```
### Map residuals

This code first reads the **Master Plan 2019 subzone boundaries** from a KML file and transforms them into Singapore’s projected coordinate system (EPSG:3414). All subzones are dissolved into a **single outline polygon of Singapore** (`sg_outline`) using `summarise()`.

Two maps are generated using the **tmap** package:

- **Map A** displays only the Singapore outline as a base map, styled with a light grey fill and black borders.
- **Map B** overlays the residuals from the OLS model (`hdb_train$residuals`) onto the Singapore map using the `HDB_train_sf` spatial dataset. Points are coloured using a diverging Red-Blue palette (`-RdBu`) centered at zero, allowing for quick visual inspection of over- and under-predicted areas in the training set.
```{r}
# 1) Read Master Plan 2019 subzones and convert to EPSG:3414
subzones <- st_read(
  "data/rawdata/MasterPlan2019SubzoneBoundaryNoSeaKML.kml",
  quiet = TRUE
) %>%
  st_transform(3414)

# 2) Make a single Singapore outline polygon (dissolve all subzones)
sg_outline <- subzones %>%
  summarise()   # dissolves into one multipart polygon

# --- A) Just the Singapore outline map ----------------------------
tmap_mode("plot")

tm_shape(sg_outline) +
  tm_polygons(col = "grey90", border.col = "black") +
  tm_layout(title = "Singapore Outline (Master Plan 2019)")

# --- B) Outline + your residual dots ------------------------------
train_resid_sf <- HDB_train_sf %>%
  mutate(residuals = hdb_train$residuals)

tm_shape(sg_outline) +
  tm_polygons(col = "grey95", border.col = "black") +
tm_shape(train_resid_sf) +
  tm_dots(col = "residuals",
          palette = "-RdBu",
          midpoint = 0,
          size = 0.23) +
  tm_layout(title = "OLS Residuals (Train Set) with Singapore Outline")
```
The residuals map overlays the OLS model’s training-set residuals onto the Singapore outline to reveal where the model systematically over- or under-predicts HDB resale prices. Each point represents an individual transaction, coloured by the magnitude and direction of its residual.

Blue tones indicate negative residuals, meaning the model predicted prices higher than the actual transacted values. These typically appear in clusters across parts of the north, west, and eastern regions, suggesting local market characteristics—such as older stock, lower amenity access, or neighbourhood-specific effects—not fully captured by the model.

Conversely, orange to red tones represent positive residuals, where actual prices exceeded predictions. These are concentrated in pockets of central, northeastern, and western estates, potentially reflecting unmodelled premiums such as micro-location desirability, school quality, renovation levels, or nearby upcoming developments.

The presence of spatially clustered residuals signals spatial autocorrelation, implying that HDB prices are influenced by local neighbourhood effects beyond the global OLS assumptions. This suggests that spatial models (e.g., GWR or spatial lag/error models) may better capture underlying patterns.

Overall, this residual map helps diagnose model limitations and highlights geographic areas where predictive accuracy can be improved.


