---
title: "Take Home Exercise 3"
author: "Gautamgovan"
date: "October 28, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 4
    number-sections: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Predictive Modelling of HDB Resale Prices with Geographically Weighted Methods

### Objective

Housing is an essential component of household wealth worldwide. Purchasing a home has always represented a major investment for most individuals. The price of housing is influenced by a wide range of factors. Some are macro-level, such as the overall state of the economy or the inflation rate, while others are property-specific. These can be further categorized into structural and locational factors.

Structural factors refer to the characteristics of the property itself, such as size, fittings, and tenure. Locational factors, on the other hand, relate to the neighbourhood context of the property, including proximity to childcare centres, public transport, and shopping facilities.

Traditionally, housing resale price models have been estimated using the Ordinary Least Squares (OLS) method. However, this approach does not account for spatial autocorrelation and spatial heterogeneity, which are commonly present in geographic datasets such as housing transactions. When spatial autocorrelation exists, OLS estimation of housing resale price models may produce biased, inconsistent, or inefficient results (Anselin, 1998). To address these limitations, Geographically Weighted Models (GWMs) have been introduced to more accurately calibrate both explanatory and predictive models for housing resale prices.

The objective of this take-home exercise is to develop a spatially informed model to predict HDB resale prices in Singapore, making use of appropriate geospatial analytics methods.

We will develop a model to predict HDB resale prices for the period July to September 2025, using HDB resale transaction data from July 2024 to June 2025.

## Installing and Loading R packages

- **`sf`** — Provides modern spatial data handling in R.  
  Used to create, transform, and manipulate spatial features (points, polygons), and perform operations like distance calculation, coordinate transformation, and spatial joins.

- **`spdep`** — Supports spatial dependence analysis.  
  Used to create neighbour lists and spatial weights, and compute Moran’s I to detect spatial autocorrelation in model residuals.

- **`GWmodel`** — Core package for Geographically Weighted Regression (GWR).  
  Used for selecting optimal bandwidths, calibrating GWR models, and generating spatially varying predictions.

- **`SpatialML`** — Enables Geographically Weighted Machine Learning models such as Geographically Weighted Random Forest (GRF).  
  Used for bandwidth selection, model fitting, and local prediction in GRF.

- **`tmap`** — Thematic mapping package for spatial data visualisation.  
  Used to visualise spatial layers, including Singapore outline maps and the spatial distribution of model residuals.

- **`rsample`** — Infrastructure for resampling and splitting data.  
  Used here for creating random subsets (e.g., for GWR calibration).

- **`Metrics`** — Provides standard model evaluation metrics.  
  Used to compute predictive accuracy measures such as RMSE and MAE for various models.

- **`tidyverse`** — Collection of packages for data wrangling and visualisation (e.g., `dplyr`, `ggplot2`, `readr`, `stringr`, `purrr`).  
  Used extensively for data cleaning, transformation, joining, and plotting.

- **`httr`** — Makes HTTP requests from R.  
  Used to query the OneMap API for address and postal code geocoding.

- **`jsonlite`** — Converts JSON responses into R-friendly data structures.  
  Used to parse JSON responses from OneMap into usable data frames.

- **`progress`** — Adds progress bars to loops and long-running operations.  
  Used to visually track geocoding progress during OneMap queries.

- **`ggplot2`** — Grammar of graphics for data visualisation.  
  Used to create histograms, bar charts, and model comparison plots.

- **`caret`** — Unified machine learning framework.  
  Used to train and tune the non-spatial Random Forest model (`ranger`) with cross-validation and hyperparameter control.

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, Metrics, tidyverse,httr, tidyverse, 
               jsonlite, progress,ggplot2,caret,gt)
```

## Data Preparation

### HDB Dataset

#### Selecting Date of Interest

This data set comprises of HDB resale transactions from Jan 2017 onwards. For the purpose of the hands on Exercise, the code chunk below will be used to extract a sub-set of transaction records on 5-Room flat transacted from July 2024 to September 2025.

```{r}
HDBresale <- read_csv(
  "data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "5 ROOM") %>%
  filter(month >= "2024-07" & month < "2025-10")
```

#### Geocoding HDB resale data

In this section, we will geocode HDB resale data by using SLA Onemap Geocode API.

Let us prepare a geocoding function by using the code chunk below. It aims to perform the following tasks:

Combining the block and street name into an address. The addresses will be used to geocode instead of postal code that was learned in In-class Exercise 3b, Passing the address as the searchVal in our query, Sending the query to OneMapSG search. Note: Since we don’t need all the address details, we can set getAddrDetails as ‘N’, Converting response (JSON object) to text, Saving response in text form as a dataframe, and Retain the latitude and longitude for our output.

```{r}
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address, 
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  res <- GET(base_url, query = query)
  restext<-content(res, as="text")
  
  output <- fromJSON(restext)  %>% 
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)

  return(output)
}
```

Next, the code chunk below will be used to to execute the geocoding function:

```{r}
#| eval: False
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for (i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

```{r}
#| eval: false
#| include: false
saveRDS(HDBresale, file = "data/rds/HDBresale_geocoded.rds")
```

```{r}
#| include: false
HDBresale <- readRDS("data/rds/HDBresale_geocoded.rds")
```

#### Converting `storey_range` to a Numerical Value (`storey_mid`)

The `storey_range` field in the HDB resale dataset is provided as a text range such as `"07 TO 09"` or `"13 TO 15"`. Since machine learning models require numerical inputs, we convert this range into a single numeric value that approximates the floor level of the unit.

This is done by:

1.  **Removing spaces** from the text
    -   Example: `"07 TO 09"` → `"07TO09"`
2.  **Splitting the range** into the lower and upper floor values
    -   `"07TO09"` → `"07"` and `"09"`
3.  **Calculating the midpoint** (average) of the two values
    -   \[ \frac{7 + 9}{2} = 8 \]
4.  **Creating a new variable** named `storey_mid` to store this midpoint value.

This numeric representation allows the floor level information to be used effectively in statistical and machine learning models.

5.  Remove the original storey_range column, as it is no longer needed.

```{r}
storey_mid <- function(x) {
  x %>%
    str_replace_all(" ", "") %>%    # remove spaces
    str_split("TO") %>%             # split into lower and upper floors
    sapply(\(v) mean(as.numeric(v)))  # convert to numeric and take midpoint
}

HDBresale <- HDBresale %>%
  mutate(storey_mid = storey_mid(storey_range)) %>%  # create numeric floor feature
  select(-storey_range)                              # remove the original text column

```

#### Converting `remaining_lease` to a Numeric Value (`remaining_lease_yrs`)

The `remaining_lease` field is provided as text (e.g., `"87 years 4 months"`). To use this information in modelling, we convert it into a numeric value representing the total lease remaining in **years**.

The function `convert_remaining_lease()`:

1.  Extracts the number of **years**.
2.  Extracts the number of **months** (or uses 0 if not specified).
3.  Converts months into a fraction of a year (`months / 12`).
4.  Adds the two together to get a single numeric value.

We then apply the function to create a new column `remaining_lease_yrs` and drop the original text column.

```{r}
convert_remaining_lease <- function(x) {
  years <- as.numeric(str_extract(x, "\\d+(?=\\syear)"))
  months <- str_extract(x, "(?<=years?\\s)\\d+(?=\\smonth)") %>% as.numeric()
  months[is.na(months)] <- 0
  years + months/12
}

HDBresale <- HDBresale %>%
  mutate(remaining_lease_yrs = convert_remaining_lease(remaining_lease)) %>%
  select(-remaining_lease)
```

### Proximity Analysis

In this section, we will learn how to count the number of geographic entities located within a defined distance from each HDB property.

#### Converting to an sf object

Before performing proximity analysis, it is important to ensure that:

-   input data sets must be in sf objects, and
-   they must be in similar projected coordinates systems.

In the code chunk below,

-   st_as_sf() is used to convert HDBresale tibble data.frame to sf object by using values from LONGITUDE and LATITUDE columns to form the geometry features.
-   st_transform() is then used to transform the sf object into svy21 projected coordinates system.
-   the output sf object is called HDBresale_sf. It is in sf data.frame format.

```{r}
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326) %>%
  st_transform(crs = 3414)
```

#### PreSchool Location

Next, we will locate and download Pre-Schools Location data from Singapore’s open data portal. There are both geojson and kml version. In this exercise, the kml version will be used.

Then, code chunk below will be used to import the preschool location data into R environment by using st_read() of readr package.

```{r}
preschool = st_read(
  "data/rawdata/PreSchoolsLocation.kml") %>%
  st_transform(crs= 3414)
```

Lastly, code chunk below will be used to count the number of pre-schools located within 350m of each HDB property.

```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(WITHIN_350M_PRESCHOOL = map_int(within_350m, length))
```

#### Proximity to CBD

The code chunk below computes the Euclidean (straight-line) distance between each flat and the Downtown Core (CBD) centroid using Singapore’s projected coordinate system (EPSG:3414, SVY21). The resulting columns proximity_to_cbd_m and proximity_to_cbd_km represent the flat’s distance to the CBD in meters and kilometers respectively

```{r}
# --- Define Downtown Core (CBD) coordinates (WGS84) ---
cbd_lon <- 103.85
cbd_lat <- 1.28967

# --- Create CBD point and transform to SVY21 (3414) ---
cbd_wgs84 <- st_sfc(st_point(c(cbd_lon, cbd_lat)), crs = 4326)
cbd_3414  <- st_transform(cbd_wgs84, 3414)

# --- Compute straight-line distance to CBD (meters / km) ---
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    proximity_to_cbd_m  = as.numeric(st_distance(geometry, cbd_3414)[, 1]),
    proximity_to_cbd_km = proximity_to_cbd_m / 1000
  )
```

```{r}
#| include: false
HDBresale_sf <- HDBresale_sf %>%
  select(-proximity_to_cbd_m)
```

#### Bus Stops within 350 metres

Read Bustops shp file and check CRS

```{r}
busstops <- st_read("data/rawdata/BusStop.shp", quiet = TRUE)%>%
  st_transform(crs= 3414)
st_crs(busstops)  # check the CRS
```

```{r}
# Count number of bus stops within 350 m for each HDB point
within_350m <- st_is_within_distance(HDBresale_sf, busstops, dist = 350)

# Add count column to main dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(WITHIN_350M_BUSSTOP = purrr::map_int(within_350m, length))
```

#### Elder care within 350 metres

Read Eldercare shapefile and check CRS

```{r}
eldercare <- st_read("data/rawdata/ELDERCARE.shp", quiet = TRUE)%>%
  st_transform(crs= 3414)
st_crs(eldercare)  # check the CRS
```

```{r}
within_350m <- st_is_within_distance(HDBresale_sf, eldercare, dist = 350)

#Add count column to main dataset

HDBresale_sf <- HDBresale_sf %>%
mutate(WITHIN_350M_ELDERCARE = purrr::map_int(within_350m, length))
```

#### Proximity to Good Primary Schools

For this part we have collected top 20 Primary Schools from Singapore in a CSV file along with its postal code , we will use One map API to geo-encode using postal code

The below code chunk gets the latitude and longitude of the school using postal codes

```{r}
# --- 1) Read the CSV (school_name, postal_code) ---
good_ps <- read_csv("data/rawdata/good_primary_schools.csv",
                    show_col_types = FALSE) %>%
  mutate(postal_code = str_pad(as.character(postal_code), 6, pad = "0"))
```

```{r}
#| eval: false

# --- 2) OneMap postal-code geocoder (returns SVY21 X/Y) ---
onemap_lookup <- function(pc){
  resp <- RETRY("GET", "https://onemap.gov.sg/api/common/elastic/search",
                query = list(searchVal = pc,
                             returnGeom = "Y",
                             getAddrDetails = "Y",
                             pageNum = "1"),
                quiet = TRUE)
  js <- content(resp, as = "parsed")
  if (!is.null(js$found) && js$found > 0) {
    r <- js$results[[1]]
    tibble(POSTAL = r$POSTAL,
           X = as.numeric(r$X),   # SVY21 easting
           Y = as.numeric(r$Y))   # SVY21 northing
  } else {
    tibble(POSTAL = pc, X = NA_real_, Y = NA_real_)
  }
}

# --- 3) Geocode unique postcodes 
pcs <- unique(good_ps$postal_code)
geo_tbl <- map_dfr(pcs, ~ tryCatch(onemap_lookup(.x),
                                   error = function(e) tibble(POSTAL=.x, X=NA_real_, Y=NA_real_)))

write_rds(geo_tbl, "data/rds/good_ps_postal_geocode.rds")

```

```{r}
geo_tbl <- read_rds("data/rds/good_ps_postal_geocode.rds")
# --- 4) Join back & keep geocoded rows; make sf in SVY21 (EPSG:3414) ---
good_ps_sf <- good_ps %>%
  left_join(geo_tbl, by = c("postal_code" = "POSTAL")) %>%
  filter(!is.na(X), !is.na(Y)) %>%
  st_as_sf(coords = c("X","Y"), crs = 3414)  # OneMap X/Y are SVY21
```

Now the below code chunk is used to derive proximity columns to good primary schools

```{r}
# --- 5) Proximity features  ---
# Count of good schools within 1 km
idx_1km <- st_is_within_distance(HDBresale_sf, good_ps_sf, dist = 1000)
cnt_1km <- lengths(idx_1km)

# Distance to nearest good school (km)
nearest_idx <- st_nearest_feature(HDBresale_sf, good_ps_sf)
nearest_dist_km <- as.numeric(st_distance(HDBresale_sf, good_ps_sf[nearest_idx, ], by_element = TRUE))/1000

# Add to main dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    WITHIN_1KM_GOOD_PS = cnt_1km,
    dist_good_ps_km    = nearest_dist_km
  )
```

#### Proximity to MRT/LRT stations

This code safely loads and processes MRT/LRT station geometries to calculate each HDB flat’s distance to its nearest station. It first reads the shapefile with error handling and removes invalid geometries. For any problematic shapes, it extracts a representative point using a safe function (safe_point). Next, it groups polygons by station name, merges them, and ensures each has one valid point in SVY21 coordinates. Finally, it computes straight-line distances (in kilometers) from each HDB flat to its nearest station and adds this as dist_mrt_km to the dataset.

```{r}

# 1) Try reading; if a few polygons fail, catch and skip
mrt_raw <- tryCatch(
  st_read("data/rawdata/RapidTransitSystemStation.shp", quiet = TRUE),
  error = function(e) {
    message("Some polygons failed to load; attempting partial recovery")
    st_read("data/rawdata/RapidTransitSystemStation.shp", quiet = TRUE, stringsAsFactors = FALSE)
  }
)

# 2) Drop any completely invalid geometries
mrt_valid <- mrt_raw[st_is_valid(mrt_raw), ]

# 3) For any remaining invalid shapes, use try() on point extraction
safe_point <- function(geom) {
  pt <- try(st_point_on_surface(geom), silent = TRUE)
  if (inherits(pt, "try-error")) NA else pt
}

# 4) Create one point per station safely
mrt_pts <- mrt_valid %>%
  group_by(STN_NAM_DE) %>%
  summarise(geometry = {
    # combine all parts and get a safe point
    geom <- st_combine(geometry)
    safe_point(geom)
  }, .groups = "drop") %>%
  filter(!st_is_empty(geometry))

st_crs(mrt_pts) <- 3414  # ensure CRS SVY21

# 5) Compute distance to nearest MRT/LRT (km)
nn_idx <- st_nearest_feature(HDBresale_sf, mrt_pts)
dist_mrt_km <- as.numeric(st_distance(HDBresale_sf, mrt_pts[nn_idx, ], by_element = TRUE)) / 1000

# 6) Add to HDB dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(dist_mrt_km = dist_mrt_km)

```

#### Proximity to Parks

Read ParkFacilities (GeoJSON) and ensure CRS

```{r}
parks <- st_read("data/rawdata/ParkFacilities.geojson", quiet = TRUE) %>%
  st_transform(3414)               # SVY21 metres
```

Compute distance to the nearest park (km) and add to dataset

```{r}
# Distance to nearest park facility (km)
nn_idx <- st_nearest_feature(HDBresale_sf, parks)
dist_park_km <- as.numeric(
  st_distance(HDBresale_sf, parks[nn_idx, ], by_element = TRUE)
) / 1000

# Add to main dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(dist_park_km = dist_park_km)
```

#### Proximity to Malls

Read the malls CSV (mall_name, postal_code)

```{r}
malls <- readr::read_csv("data/rawdata/singapore_malls_postal_codes.csv",
                         show_col_types = FALSE) %>%
  dplyr::mutate(postal_code = stringr::str_pad(as.character(postal_code), 6, pad = "0"))
```

Geocode mall postal codes with OneMap (SVY21 X/Y)

```{r}
#| eval: false
#| 
onemap_lookup <- function(pc){
  resp <- httr::RETRY("GET", "https://onemap.gov.sg/api/common/elastic/search",
                      query = list(searchVal = pc,
                                   returnGeom = "Y",
                                   getAddrDetails = "Y",
                                   pageNum = "1"),
                      quiet = TRUE)
  js <- httr::content(resp, as = "parsed")
  if (!is.null(js$found) && js$found > 0) {
    r <- js$results[[1]]
    tibble::tibble(POSTAL = r$POSTAL,
                   X = as.numeric(r$X),   # SVY21 easting
                   Y = as.numeric(r$Y))   # SVY21 northing
  } else {
    tibble::tibble(POSTAL = pc, X = NA_real_, Y = NA_real_)
  }
}

pcs <- unique(malls$postal_code)
mall_geo_tbl <- purrr::map_dfr(
  pcs,
  ~ tryCatch(onemap_lookup(.x),
             error = function(e) tibble::tibble(POSTAL=.x, X=NA_real_, Y=NA_real_))
)

readr::write_rds(mall_geo_tbl, "data/rds/malls_postal_geocode.rds")
```

Join geocodes and create an sf layer (SVY21 / EPSG:3414)

```{r}
mall_geo_tbl <- readr::read_rds("data/rds/malls_postal_geocode.rds")

malls_sf <- malls %>%
  dplyr::left_join(mall_geo_tbl, by = c("postal_code" = "POSTAL")) %>%
  dplyr::filter(!is.na(X), !is.na(Y)) %>%
  sf::st_as_sf(coords = c("X","Y"), crs = 3414)
```

Derive proximity features to malls

```{r}
# Distance to nearest mall (km)
mall_nn_idx <- sf::st_nearest_feature(HDBresale_sf, malls_sf)
dist_mall_km <- as.numeric(sf::st_distance(HDBresale_sf, malls_sf[mall_nn_idx, ], by_element = TRUE)) / 1000

# Count of malls within 1 km (adjust if you prefer another radius)
malls_within_1km <- sf::st_is_within_distance(HDBresale_sf, malls_sf, dist = 1000)
cnt_malls_1km <- lengths(malls_within_1km)

# Add to main dataset
HDBresale_sf <- HDBresale_sf %>%
  dplyr::mutate(
    dist_mall_km       = dist_mall_km,
    WITHIN_1KM_MALLS   = cnt_malls_1km
  )
```

#### Proximity to CHAS clinics

Read Chas data and ensure CRS

```{r}
chas <- st_read("data/rawdata/CHASClinics.geojson", quiet = TRUE) %>%
  st_transform(3414)
```

```{r}
# Distance to nearest CHAS clinic (km)
chas_nn_idx <- st_nearest_feature(HDBresale_sf, chas)
dist_chas_km <- as.numeric(st_distance(HDBresale_sf, chas[chas_nn_idx, ], by_element = TRUE)) / 1000

# Count of CHAS clinics within 350 m
chas_within_350m <- st_is_within_distance(HDBresale_sf, chas, dist = 350)
cnt_chas_350m <- lengths(chas_within_350m)

# Add to main dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    dist_chas_km    = dist_chas_km,
    WITHIN_350M_CHAS = cnt_chas_350m
  )

```

#### Proximity to Hawker Centres

Read Hawker Centres GeoJSON and check CRS

```{r}
hawker <- st_read("data/rawdata/HawkerCentresGEOJSON.geojson", quiet = TRUE)%>%
  st_transform(3414)

st_crs(hawker)
```

Compute distance to nearest hawker centre (in km) ---

```{r}
hawker_nn_idx <- st_nearest_feature(HDBresale_sf, hawker)
dist_hawker_km <- as.numeric(st_distance(HDBresale_sf, hawker[hawker_nn_idx, ], by_element = TRUE)) / 1000

#  Add distance feature to HDB resale dataset
HDBresale_sf <- HDBresale_sf %>%
  mutate(dist_hawker_km = dist_hawker_km)
```

#### Childcare Centres within 350 metres

Read ChildCare Services (GeoJSON) and ensure SVY21:

```{r}
childcare <- st_read("data/rawdata/ChildCareServices.geojson", quiet = TRUE)%>%st_transform(3414)

st_crs(hawker)

```

Compute count within 350 m and add to main data

```{r}
within_350m_cc <- st_is_within_distance(HDBresale_sf, childcare, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(WITHIN_350M_CHILDCARE = lengths(within_350m_cc))
```

```{r}
#| eval: false
#| include: false
colnames(HDBresale_sf)
```

### Calculate Correlation Matrix

Before loading the predictors into a predictive model, it is always a good practice to use correlation matrix to examine if there is sign of multicollinearity.

```{r}
# 1) Drop geometry
HDBresale_nogeo <- HDBresale_sf %>%
  st_drop_geometry()

# 2) Select only numeric columns for correlation
numeric_cols <- HDBresale_nogeo %>%
  select(where(is.numeric))

# 3) Compute correlation matrix
corr_matrix <- cor(numeric_cols, use = "complete.obs")

# 4) Plot correlation matrix
corrplot::corrplot(corr_matrix,
                   diag = FALSE,
                   order = "AOE",
                   tl.pos = "td",
                   tl.cex = 0.6,
                   method = "number",
                   type = "upper")
```

Drop Highly correlated Values

```{r}
HDBresale_sf <- HDBresale_sf %>%
  select(-WITHIN_350M_PRESCHOOL,
         -lease_commence_date,
         -WITHIN_1KM_MALLS,
         -WITHIN_1KM_GOOD_PS)
```

### Train–Test Split for HDB Resale Modelling

This code chunk separates the dataset into **training** and **testing** periods and removes geometry for modelling.

1.  Training Set (Jul 2024 – Jun 2025) Filters `HDBresale_sf` to include resale transactions from July 2024 to June 2025, storing it as `HDB_train_sf`.

2.  Test Set (Jul 2025 – Sep 2025) Filters the same dataset for resale transactions from July 2025 to September 2025, creating `HDB_test_sf`.

3.  Drop Geometry for Modelling Converts both `sf` objects into regular data frames (`hdb_train`, `hdb_test`) by removing spatial geometry columns using `st_drop_geometry()`, preparing them for statistical or machine-learning models.

```{r}

# Training: Jul 2024 – Jun 2025
HDB_train_sf <- HDBresale_sf %>%
  dplyr::filter(month >= "2024-07" & month <= "2025-06")

# Test (prediction period): Jul 2025 – Sep 2025
HDB_test_sf <- HDBresale_sf %>%
  dplyr::filter(month >= "2025-07" & month <= "2025-09")

# For modelling (no geometry)
hdb_train <- HDB_train_sf %>% sf::st_drop_geometry()
hdb_test  <- HDB_test_sf  %>% sf::st_drop_geometry()
```

```{r}

ggplot(hdb_train, aes(x = resale_price)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of HDB Resale Prices (Raw)",
    x = "Resale Price (SGD)",
    y = "Count"
  ) +
  theme_minimal()
```

## Building a non-spatial multiple linear regression

This code chunk fits a linear regression model (`lm_ols`) using the `lm()` function, where the log-transformed HDB resale price is the dependent variable. The model includes structural attributes (e.g., floor area, storey level, remaining lease, flat model) and locational factors such as proximity to CBD, MRT stations, parks, malls, and amenities like bus stops, childcare, eldercare, and CHAS clinics. The independent variables are selected based on their potential influence on resale price. The model is trained on the `hdb_train` dataset.

```{r}
lm_ols <- lm(
  log(resale_price) ~ 
    floor_area_sqm +
    storey_mid +
    remaining_lease_yrs +
    flat_model +
    proximity_to_cbd_km +
    WITHIN_350M_BUSSTOP +
    WITHIN_350M_ELDERCARE +
    dist_good_ps_km +
    dist_mrt_km +
    dist_park_km +
    dist_mall_km +
    dist_hawker_km +
    WITHIN_350M_CHILDCARE +
    WITHIN_350M_CHAS,
  data = hdb_train
)
```

### OLS Residual Diagnostics & Spatial Dependence Check

#### Summary of the model

```{r}
summary(lm_ols)
```

The OLS model explains resale prices well, with strong effects from floor area, storey height, lease, and proximity to key amenities. Significant spatial patterns remain, motivating geographically weighted modelling.

```{r}
# Add residuals and run standard OLS diagnostics
hdb_train$residuals <- residuals(lm_ols)

par(mfrow = c(2, 2))
plot(lm_ols)   # includes Residuals vs Fitted, Q-Q plot, etc.

# Check multicollinearity
car::vif(lm_ols)
```

These OLS diagnostic plots show several issues with model assumptions. The Residuals vs Fitted plot indicates slight non-linearity and heteroscedasticity. The Q-Q plot shows heavy tails, meaning residuals are not normally distributed. The Scale-Location plot confirms non-constant variance, while the Residuals vs Leverage plot shows a few high-leverage points that may disproportionately influence the model.

#### Spatial autocorrelation of residuals (Moran’s I)

```{r}

# Build neighbour structure and spatial weights
coords <- st_coordinates(HDB_train_sf)
nb     <- knn2nb(knearneigh(coords, k = 8))
lw     <- nb2listw(nb, style = "W")

# Test for spatial autocorrelation in OLS residuals
moran.test(hdb_train$residuals, lw)
```

OLS residuals exhibit strong and significant spatial autocorrelation (Moran’s I = 0.668, p \< 0.001), indicating substantial spatial heterogeneity and justifying the use of geographically weighted models.

### Predict on the test set (Jul–Sep 2025)

```{r}
hdb_test$pred_log <- predict(lm_ols, newdata = hdb_test)
hdb_test$pred_price <- exp(hdb_test$pred_log)
```

### Evaluate predictive performance

```{r}
library(Metrics)

rmse_ols <- rmse(hdb_test$resale_price, hdb_test$pred_price)
mae_ols  <- mae(hdb_test$resale_price, hdb_test$pred_price)
r2_ols   <- 1 - sum((hdb_test$resale_price - hdb_test$pred_price)^2) /
                 sum((hdb_test$resale_price - mean(hdb_test$resale_price))^2)

rmse_ols
mae_ols
r2_ols
```

### Map residuals

This code first reads the **Master Plan 2019 subzone boundaries** from a KML file and transforms them into Singapore’s projected coordinate system (EPSG:3414). All subzones are dissolved into a **single outline polygon of Singapore** (`sg_outline`) using `summarise()`.

Two maps are generated using the **tmap** package:

-   **Map A** displays only the Singapore outline as a base map, styled with a light grey fill and black borders.
-   **Map B** overlays the residuals from the OLS model (`hdb_train$residuals`) onto the Singapore map using the `HDB_train_sf` spatial dataset. Points are coloured using a diverging Red-Blue palette (`-RdBu`) centered at zero, allowing for quick visual inspection of over- and under-predicted areas in the training set.

```{r}
# 1) Read Master Plan 2019 subzones and convert to EPSG:3414
subzones <- st_read(
  "data/rawdata/MasterPlan2019SubzoneBoundaryNoSeaKML.kml",
  quiet = TRUE
) %>%
  st_transform(3414)

# 2) Make a single Singapore outline polygon (dissolve all subzones)
sg_outline <- subzones %>%
  summarise()   # dissolves into one multipart polygon

# --- A) Just the Singapore outline map ----------------------------
tmap_mode("plot")

tm_shape(sg_outline) +
  tm_polygons(col = "grey90", border.col = "black") +
  tm_layout(title = "Singapore Outline (Master Plan 2019)")

# --- B) Outline + residual dots ------------------------------
train_resid_sf <- HDB_train_sf %>%
  mutate(residuals = hdb_train$residuals)

tm_shape(sg_outline) +
  tm_polygons(col = "grey95", border.col = "black") +
tm_shape(train_resid_sf) +
  tm_dots(col = "residuals",
          palette = "-RdBu",
          midpoint = 0,
          size = 0.23) +
  tm_layout(title = "OLS Residuals (Train Set) with Singapore Outline")
```

The residuals map overlays the OLS model’s training-set residuals onto the Singapore outline to reveal where the model systematically over- or under-predicts HDB resale prices. Each point represents an individual transaction, coloured by the magnitude and direction of its residual.

Blue tones indicate negative residuals, meaning the model predicted prices higher than the actual transacted values. These typically appear in clusters across parts of the north, west, and eastern regions, suggesting local market characteristics—such as older stock, lower amenity access, or neighbourhood-specific effects—not fully captured by the model.

Conversely, orange to red tones represent positive residuals, where actual prices exceeded predictions. These are concentrated in pockets of central, northeastern, and western estates, potentially reflecting unmodelled premiums such as micro-location desirability, school quality, renovation levels, or nearby upcoming developments.

The presence of spatially clustered residuals signals spatial autocorrelation, implying that HDB prices are influenced by local neighbourhood effects beyond the global OLS assumptions. This suggests that spatial models (e.g., GWR or spatial lag/error models) may better capture underlying patterns.

Overall, this residual map helps diagnose model limitations and highlights geographic areas where predictive accuracy can be improved.

## GWR Modelling

### Data prep

Randomly sample up to 5,000 training points

We limit the GWR training sample to 5,000 points because GWR is computationally heavy—it fits a local regression at every data point, making runtime and memory usage grow rapidly with larger datasets. A 5,000-point sample provides sufficient spatial coverage across Singapore while keeping bandwidth selection and model fitting fast, stable, and computationally feasible.

```{r}
set.seed(123)  # reproducible sampling

n_target <- 5000   # maximum allowed size for GWR

if (nrow(HDB_train_sf) > n_target) {
HDB_train_sf_sub <- HDB_train_sf %>%
dplyr::slice_sample(n = n_target)
} else {
HDB_train_sf_sub <- HDB_train_sf
}

nrow(HDB_train_sf_sub)  # check actual sample size
```

### Select optimal GWR bandwidth

This code uses bw.gwr() to automatically select the optimal bandwidth for the GWR model. The formula specifies resale price and all predictor variables, while data = HDB_train_sp_sub supplies the sampled training points. A Gaussian kernel and adaptive bandwidth are used, meaning each location uses a variable number of neighbours. longlat = FALSE ensures distances are computed in the SVY21 projected coordinate system. The resulting bw_gwr is the best-fitting bandwidth for the GWR model.

```{r}
#| eval: false

bw_gwr <- bw.gwr(
formula = resale_price ~
floor_area_sqm + storey_mid + remaining_lease_yrs + flat_model +
proximity_to_cbd_km + WITHIN_350M_BUSSTOP + WITHIN_350M_ELDERCARE +
dist_good_ps_km + dist_mrt_km + dist_park_km + dist_mall_km +
dist_hawker_km + WITHIN_350M_CHILDCARE + WITHIN_350M_CHAS,
data = HDB_train_sp_sub,
kernel = 'gaussian',
adaptive = TRUE,
longlat = FALSE
)
bw_gwr
```

```{r}
#| eval: false
#| include: false
saveRDS(bw_gwr, file = "data/rds/bw_gwr.rds")
```

```{r}
#| include: false
bw_gwr <- readRDS("data/rds/bw_gwr.rds")
```

### Fit GWR model on sampled training set

```{r}
#| eval: false
gwr_fit <- gwr.basic(
formula  = resale_price ~
floor_area_sqm + storey_mid + remaining_lease_yrs + flat_model +
proximity_to_cbd_km + WITHIN_350M_BUSSTOP + WITHIN_350M_ELDERCARE +
dist_good_ps_km + dist_mrt_km + dist_park_km + dist_mall_km +
dist_hawker_km + WITHIN_350M_CHILDCARE + WITHIN_350M_CHAS,
data     = HDB_train_sp_sub,
bw       = bw_gwr,
kernel   = "gaussian",
adaptive = TRUE,
longlat  = FALSE
)


# Save GWR model
saveRDS(gwr_fit, file = "data/rds/gwr_fit.rds")

```

```{r}
gwr_fit <- readRDS("data/rds/gwr_fit.rds")

# Diagnostics
gwr_fit$GW.diagnostic
```

### Use GWR Model to Predict on Test Data

#### Convert test set to Spatial

```{r}
HDB_test_sp <- as(HDB_test_sf, "Spatial")
```

#### Predict using the fitted GWR model (gwr_fit)

In code chunks below - A GWR formula (`gwr_formula`) is defined with `resale_price` as the dependent variable and multiple structural and locational predictors on the right-hand side. - Variables include attributes like `floor_area_sqm`, `storey_mid`, and distances to amenities such as `dist_mrt_km`, `dist_park_km`, and `WITHIN_350M_CHILDCARE`. - `gwr_vars` extracts all variable names on the RHS of the formula using `all.vars()` and excludes the dependent variable using `[-1]`. - `test_cols` retrieves the column names of the `HDB_test_sp@data` SpatialPointsDataFrame, which will be used to apply the trained GWR model. - The `setdiff()` function compares `gwr_vars` against `test_cols` to find any missing predictor columns in the test set. - This acts as a **safeguard** to ensure the test data contains all required variables for prediction, avoiding runtime errors in GWR modelling.

```{r}
gwr_formula <- resale_price ~
  floor_area_sqm + storey_mid + remaining_lease_yrs +
  proximity_to_cbd_km + WITHIN_350M_BUSSTOP + WITHIN_350M_ELDERCARE +
  dist_good_ps_km + dist_mrt_km + dist_park_km + dist_mall_km +
  dist_hawker_km + WITHIN_350M_CHILDCARE + WITHIN_350M_CHAS

```

```{r}
# variables used on the RHS of the formula
gwr_vars <- all.vars(gwr_formula)[-1]   # drop response

# columns in your test SpatialPointsDataFrame
test_cols <- names(HDB_test_sp@data)

setdiff(gwr_vars, test_cols)
```

The code chunk below does the following - `vars_all` stores all variables required for GWR: both the response (`resale_price`) and all predictors from `gwr_vars`. - The `@data` slot of the `HDB_test_sp` SpatialPointsDataFrame is extracted and a `row_id` is added to preserve row positions after filtering. - `test_df` keeps only the necessary columns (`row_id` + `vars_all`) to be used for prediction. - `test_df_clean` filters out any rows that contain missing values in the selected columns using `complete.cases()`. - The `valid_rows` vector stores row indices of complete observations and is used to subset `HDB_test_sp` into `HDB_test_sp_clean`, which is safe for GWR prediction.

-   The second chunk (set to not evaluate with `eval: false`) calls `gwr.predict()` to make predictions on the cleaned test spatial data, using the trained model from `HDB_train_sp_sub` and a bandwidth `bw_gwr`.
-   The predicted prices are stored in a new column `gwr_pred_price` in the original `hdb_test` dataframe.
-   Only the rows corresponding to `valid_rows` are populated with predicted values to maintain alignment with the full test dataset.

```{r}
# All variables needed for GWR (response + predictors)
vars_all <- c("resale_price", gwr_vars)

# Work on the data slot of HDB_test_sp
test_df <- HDB_test_sp@data %>%
  mutate(row_id = row_number()) %>%
  select(row_id, all_of(vars_all))

# Keep only complete rows
test_df_clean <- test_df %>%
  filter(complete.cases(across(all_of(vars_all))))

valid_rows <- test_df_clean$row_id

# Subset SpatialPointsDataFrame to these rows
HDB_test_sp_clean <- HDB_test_sp[valid_rows, ]
```

-   This code chunk (set to not evaluate with `eval: false`) calls `gwr.predict()` to make predictions on the cleaned test spatial data, using the trained model from `HDB_train_sp_sub` and a bandwidth `bw_gwr`.
-   The predicted prices are stored in a new column `gwr_pred_price` in the original `hdb_test` dataframe.
-   Only the rows corresponding to `valid_rows` are populated with predicted values to maintain alignment with the full test dataset.

```{r}
#| eval: false
gwr_pred <- gwr.predict(
  formula     = gwr_formula,
  data        = HDB_train_sp_sub,
  predictdata = HDB_test_sp_clean,
  bw          = bw_gwr,
  kernel      = "gaussian",
  adaptive    = TRUE,
  longlat     = FALSE
)

# Add predictions back to hdb_test
hdb_test$gwr_pred_price <- NA_real_
hdb_test$gwr_pred_price[valid_rows] <- gwr_pred$SDF$prediction
```

```{r}
nrow(HDB_test_sp)        # original test size
nrow(HDB_test_sp_clean)  # rows used for GWR prediction
```

```{r}
#| eval: false
#| include: false
saveRDS(gwr_pred, file = "data/rds/gwr_pred.rds")
saveRDS(hdb_test, "data/rds/hdb_test_with_gwr_pred.rds")
```

```{r}
gwr_pred <- readRDS("data/rds/gwr_pred.rds")
HDB_test_results <- readRDS("data/rds/hdb_test_with_gwr_pred.rds")
```

### Evaulate the model

```{r}
library(Metrics)

idx <- !is.na(HDB_test_results$gwr_pred_price)

rmse_gwr <- rmse(HDB_test_results$resale_price[idx],
                 HDB_test_results$gwr_pred_price[idx])

mae_gwr  <- mae(HDB_test_results$resale_price[idx],
                HDB_test_results$gwr_pred_price[idx])

r2_gwr   <- 1 - sum((HDB_test_results$resale_price[idx] -
                     HDB_test_results$gwr_pred_price[idx])^2) /
                 sum((HDB_test_results$resale_price[idx] -
                      mean(HDB_test_results$resale_price[idx]))^2)

rmse_gwr
mae_gwr
r2_gwr
```

### Compute Moran’s I for GWR residuals

```{r}
HDB_test_results_sf <- HDB_test_sf %>%
  mutate(
    gwr_pred_price = HDB_test_results$gwr_pred_price,
    gwr_residual   = resale_price - gwr_pred_price
  )

coords_test <- st_coordinates(HDB_test_results_sf)
nb_test <- knn2nb(knearneigh(coords_test, k = 8))
lw_test <- nb2listw(nb_test, style = "W")

moran.test(HDB_test_results_sf$gwr_residual, lw_test)
```

The Moran’s I statistic for the GWR residuals is 0.227 (p \< 0.001), substantially lower than the OLS residual Moran’s I of approximately 0.668. This indicates that GWR captures a significant portion of the spatial structure in HDB resale prices that the global OLS model fails to account for. While some positive spatial autocorrelation remains, the large reduction in Moran’s I demonstrates that geographically varying relationships—such as differing effects of proximity to amenities and structural attributes across neighbourhoods—play an important role in price formation. Therefore, GWR provides a more spatially appropriate residual pattern and better represents local market heterogeneity compared to OLS.

## Non Spatial Random Forest

### Define predictors

```{r}
rf_formula <- resale_price ~
floor_area_sqm +
storey_mid +
remaining_lease_yrs +
flat_model +
proximity_to_cbd_km +
WITHIN_350M_BUSSTOP +
WITHIN_350M_ELDERCARE +
dist_good_ps_km +
dist_mrt_km +
dist_park_km +
dist_mall_km +
dist_hawker_km +
WITHIN_350M_CHILDCARE +
WITHIN_350M_CHAS
```

### Set up cross-validation and tuning grid

This code chunk prepares the Random Forest model for training and hyperparameter tuning.  
`trainControl()` sets up 5-fold cross-validation, ensuring the model is evaluated on five different train–validation splits for reliable performance.  
The `expand.grid()` block defines the hyperparameter search space:  
- `mtry` controls how many predictors are considered at each split,  
- `splitrule = "variance"` specifies the regression splitting criterion, and  
- `min.node.size` sets the minimum number of samples allowed in terminal nodes.  
These combinations are systematically tested to identify the best-performing Random Forest model.


```{r}
set.seed(123)

#5-fold cross-validation

rf_ctrl <- trainControl(
method = "cv",
number = 5,
verboseIter = FALSE
)

#Hyperparameter grid for calibration

rf_grid <- expand.grid(
mtry = c(3, 5, 7, 9), # number of variables tried at each split
splitrule = "variance", # regression split rule
min.node.size = c(5, 10, 20) # minimum node size
)
```

### Train and calibrate the Random Forest

```{r}
#| eval: false
set.seed(123)

rf_model <- train(
rf_formula,
data = hdb_train,
method = "ranger",
trControl = rf_ctrl,
tuneGrid = rf_grid,
importance = "permutation",
num.trees = 500
)

rf_model
rf_model$bestTune
```

```{r}
#| eval: false
#| include: false
saveRDS(rf_model, file = "data/rds/rf_model.rds")
```

```{r}
#| include: false
rf_model <- readRDS("data/rds/rf_model.rds")
```

### Predict on the test set (Jul–Sep 2025)

```{r}
hdb_test$rf_pred_price <- predict(rf_model, newdata = hdb_test)
```

### Evaluate Random Forest predictive performance

```{r}
rmse_rf <- rmse(hdb_test$resale_price, hdb_test$rf_pred_price)
mae_rf <- mae(hdb_test$resale_price, hdb_test$rf_pred_price)
r2_rf <- 1 - sum((hdb_test$resale_price - hdb_test$rf_pred_price)^2) /
sum((hdb_test$resale_price - mean(hdb_test$resale_price))^2)

rmse_rf
mae_rf
r2_rf
```

## Geographically Weighted Random Forest (GRF)

### Response + predictor variables from the RF formula

```{r}
grf_vars <- all.vars(rf_formula)
resp_var <- grf_vars[1] # "resale_price"
pred_vars <- grf_vars[-1]
```

### Training dataframe for GRF (no geometry)

```{r}
hdb_train_grf <- HDB_train_sf_sub %>%
sf::st_drop_geometry() %>%
dplyr::select(all_of(c(resp_var, pred_vars)))
```

### Coordinates matrix (X, Y) for GRF

```{r}
coords_grf <- sf::st_coordinates(HDB_train_sf_sub)

nrow(hdb_train_grf)
head(hdb_train_grf)
```

### Select optimal GRF bandwidth

```{r}
#| eval: false
set.seed(123)
grf_bw <- grf.bw(
formula = rf_formula,
dataset = hdb_train_grf,
coords = coords_grf,
kernel = "adaptive",
bw.min = 50,
bw.max = 400,
step = 25,
trees = 500,
importance = "impurity",
nthreads = parallel::detectCores() - 1,
forests = FALSE,
geo.weighted = TRUE
)

grf_bw$best.bw
```

```{r}
#| eval: false
#| include: false
best_bw_grf <- 375
saveRDS(grf_bw, file = "data/rds/grf_bw.rds")

```

```{r}
best_bw_grf <- readRDS("data/rds/grf_bw.rds")
best_bw_grf

```

### Fit GRF model on the training set

Using the optimal bandwidth, we calibrate the geographically weighted random forest. The fitted object contains both a global RF and a set of local RF models, plus local goodness-of-fit and variable importance metrics.

```{r}
#| eval: false
set.seed(123)

grf_fit <- grf(
formula = rf_formula,
dframe = hdb_train_grf,
bw = 375,
kernel = "adaptive",
coords = coords_grf,
ntree = 500,
mtry = NULL, # default p/3
importance = "impurity",
nthreads = parallel::detectCores() - 1,
forests = TRUE, # TRUE if you want all local forests
geo.weighted = TRUE,
print.results = TRUE
)

saveRDS(grf_fit, "data/rds/grf_fit.rds")
```

```{r}
#| eval: false
grf_fit <- readRDS("data/rds/grf_fit.rds")
```

Quick look at local goodness-of-fit summary

```{r}
#| eval: false
class(grf_fit) <- c("GRF", "list")
```

```{r}
#| eval: false
class(grf_fit)


```

```{r}
#| eval: false

str(grf_fit$LGofFit)
head(grf_fit$lModelSummary)
```

### Predict with GRF on the test set (Jul–Sep 2025)

We now use predict.grf() to obtain spatially varying predictions on the test period, using the nearest local RF model for each test location.

#### Test predictors and coordinates

```{r}
#| eval: false
# 1. Drop geometry but keep all predictors
test_data_nogeom <- HDB_test_sf %>%
  sf::st_drop_geometry()

# 2. Add X / Y columns from the sf geometry
coords_test <- sf::st_coordinates(HDB_test_sf)
test_data_nogeom$X <- coords_test[, 1]
test_data_nogeom$Y <- coords_test[, 2]

```

#### GRF predictions (resale_price)

```{r}
#| eval: false
grf_pred <- predict.grf(
  grf_fit,
  test_data_nogeom,
  x.var.name = "X",
  y.var.name = "Y",
  local.w    = 1,
  global.w   = 0
)
```

```{r}
#| eval: false
HDB_test_results_grf <- hdb_test %>%
  dplyr::mutate(grf_pred_price = as.numeric(grf_pred))
```

```{r}
#| eval: false
#| include: false
readr::write_rds(HDB_test_results_grf, "data/rds/HDB_test_results_grf.rds")
```

```{r}
HDB_test_results_grf <- readr::read_rds("data/rds/HDB_test_results_grf.rds")
```

### Metrics

```{r}
rmse_grf <- rmse(HDB_test_results_grf$resale_price,
HDB_test_results_grf$grf_pred_price)

mae_grf <- mae(HDB_test_results_grf$resale_price,
HDB_test_results_grf$grf_pred_price)

r2_grf <- 1 - sum((HDB_test_results_grf$resale_price -
HDB_test_results_grf$grf_pred_price)^2) /
sum((HDB_test_results_grf$resale_price -
mean(HDB_test_results_grf$resale_price))^2)

rmse_grf
mae_grf
r2_grf
```

## Comparison of models

### Summary table of model performance
This code  chunk creates a summary table comparing the performance of all four models.  
A tibble (`model_metrics`) stores RMSE, MAE, and R² values for OLS, GWR, Random Forest, and Geographically Weighted RF.  
The table is then sorted by RMSE and formatted using `gt()`, which rounds the numeric values and adds a title and subtitle.  
The result is a clean, publication-ready performance summary table suitable for reporting and analysis.


```{r}
model_metrics <- tibble::tibble(
Model = c("OLS", "GWR", "Random Forest", "Geog. Weighted RF"),
RMSE = c(rmse_ols, rmse_gwr, rmse_rf, rmse_grf),
MAE = c(mae_ols, mae_gwr, mae_rf, mae_grf),
R2 = c(r2_ols, r2_gwr, r2_rf, r2_grf)
)

model_metrics %>%
arrange(RMSE) %>%
gt() %>%
fmt_number(columns = c(RMSE, MAE), decimals = 0) %>%
fmt_number(columns = R2, decimals = 3) %>%
tab_header(title = "Predictive Performance Across Models",
subtitle = "Test period: Jul–Sep 2025"
)
```
### Bar chart comparison of RMSE / MAE / R²
This code chunk reshapes the model performance table into a long format using `pivot_longer()`, so each metric (RMSE, MAE, R²) becomes a separate row.  
A bar chart is then created using `ggplot2` to display model performance side-by-side.  
`facet_wrap()` creates individual panels for each metric, allowing easy comparison across models.  
`geom_text()` overlays the numeric values on top of each bar to enhance readability.  
The plot is styled with a minimal theme, angled x-axis labels for clarity, and bold facet titles.  
This produces a clean and visually effective comparison of model performance.

```{r}
#| label: bar chart
metrics_long <- model_metrics %>%
tidyr::pivot_longer(
cols = c(RMSE, MAE, R2),
names_to = "Metric",
values_to = "Value"
)

ggplot(metrics_long, aes(x = Model, y = Value)) +
geom_col() +
facet_wrap(~ Metric, scales = "free_y") +
geom_text(aes(label = round(Value, ifelse(Metric == "R2", 3, 0))),
vjust = -0.4, size = 3) +
labs(
title = "Model Performance Comparison",
x = NULL,
y = NULL
) +
theme_minimal(base_size = 11) +
theme(
strip.text = element_text(face = "bold"),
axis.text.x = element_text(angle = 20, hjust = 1)
)
```
The bar graph compares the out-of-sample performance of four models on HDB resale prices for Jul–Sep 2025. Random Forest (RF) is the best-performing model, achieving the lowest RMSE (48,468) and lowest MAE (33,798), with the highest R² (0.923). This indicates that a non-spatial, nonlinear ensemble can already capture most of the variation in resale prices by flexibly modelling interactions between structural and locational variables.

The Geographically Weighted Regression (GWR) model performs second best, with RMSE 56,689, MAE 40,019, and R² 0.894. GWR improves clearly over the global OLS model, confirming that allowing coefficients to vary across space helps capture local housing market differences, even within a linear framework.

The OLS model yields RMSE 81,416, MAE 58,802, and R² 0.781. Although it provides a reasonable baseline, its lower R² and higher errors reflect the limitations of a single global linear relationship in the presence of spatial heterogeneity and nonlinear effects.

Finally, the Geographically Weighted Random Forest (GRF) performs the weakest (RMSE 123,421, MAE 94,785, R² 0.498) in this implementation. This suggests that the added model complexity and local forest fitting did not translate into better generalisation on the test period—likely due to bandwidth choice, overfitting to local noise, or the relatively limited sample size for each local model.

Overall, these results indicate that a well-tuned global Random Forest provides the most accurate and stable predictions, while GWR offers a strong spatially explicit alternative that still outperforms traditional OLS.


### Actual vs predicted scatter by model
This code  chunk creates a scatterplot comparing actual and predicted resale prices for all four models.  
A tibble (`compare_df`) is first constructed containing the actual prices and each model’s predictions.  
`pivot_longer()` converts the data into long format so each model appears as a separate row.  
`ggplot()` then plots actual vs. predicted prices, adds a 45-degree reference line to indicate perfect prediction,  
and uses `facet_wrap()` to display one panel per model.  
The result visually shows how closely each model's predictions match the actual resale prices.

```{r}
compare_df <- tibble::tibble(
resale_price = HDB_test_results_grf$resale_price,
OLS = HDB_test_results$pred_price,
GWR = HDB_test_results$gwr_pred_price,
`Random Forest` = hdb_test$rf_pred_price,
`Geog. Weighted RF` = HDB_test_results_grf$grf_pred_price
)

compare_long <- compare_df %>%
tidyr::pivot_longer(
cols = -resale_price,
names_to = "Model",
values_to = "pred_price"
) %>%
filter(!is.na(pred_price))

ggplot(compare_long,
aes(x = resale_price, y = pred_price)) +
geom_point(alpha = 0.4) +
geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
facet_wrap(~ Model, scales = "free") +
scale_x_continuous(labels = scales::comma) +
scale_y_continuous(labels = scales::comma) +
labs(
title = "Actual vs Predicted HDB Resale Prices",
x = "Actual price (SGD)",
y = "Predicted price (SGD)"
) +
theme_minimal(base_size = 11)
```
The scatterplots provide a visual assessment of how well each model predicts resale prices for the Jul–Sep 2025 test period. The 45-degree dashed line represents perfect prediction, where predicted values exactly match actual transaction prices. Deviations from this line indicate under- or over-prediction.

The Random Forest model exhibits the tightest clustering around the reference line, reflecting its strong predictive performance. Most points lie very close to the 45-degree line across the full price range (approximately $600,000 to $1.6 million), indicating that the RF model captures nonlinear effects and complex interactions effectively. This visual pattern aligns with its top performance metrics (lowest RMSE and highest R²).

The GWR model also shows relatively close alignment to the reference line, though with slightly wider dispersion than Random Forest. While GWR captures spatial heterogeneity well, its linear structure limits its ability to fully model nonlinear price dynamics, leading to moderate under- and over-prediction in mid- and high-priced units.

The OLS model shows noticeably larger spread around the reference line, indicating weaker predictive accuracy. High-priced transactions above $1.2 million tend to be systematically under-predicted, suggesting that a global linear specification does not accommodate varying relationships across neighbourhoods or price tiers. This visual pattern aligns with OLS having the third-best R² and higher errors.

The Geographically Weighted Random Forest (GRF) shows the weakest fit, with a wide horizontal spread and clear bias in the upper price ranges where predictions flatten. This suggests that the GRF configuration used may be overfitting to local noise or using suboptimal bandwidth settings, resulting in poorer generalisation.

Overall, the scatterplots reinforce the quantitative results: Random Forest provides the most accurate and stable predictions, while GWR improves over OLS but remains limited by linearity. GRF underperforms in this implementation, indicating that additional tuning may be required for effective spatial machine learning.



