---
title: "Take Home Exercise 3"
author: "Gautamgovan"
date: "October 28, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 4
    number-sections: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Predictive Modelling of HDB Resale Prices with Geographically Weighted Methods

### Objective

Housing is an essential component of household wealth worldwide. Purchasing a home has always represented a major investment for most individuals. The price of housing is influenced by a wide range of factors. Some are macro-level, such as the overall state of the economy or the inflation rate, while others are property-specific. These can be further categorized into structural and locational factors.

Structural factors refer to the characteristics of the property itself, such as size, fittings, and tenure. Locational factors, on the other hand, relate to the neighbourhood context of the property, including proximity to childcare centres, public transport, and shopping facilities.

Traditionally, housing resale price models have been estimated using the Ordinary Least Squares (OLS) method. However, this approach does not account for spatial autocorrelation and spatial heterogeneity, which are commonly present in geographic datasets such as housing transactions. When spatial autocorrelation exists, OLS estimation of housing resale price models may produce biased, inconsistent, or inefficient results (Anselin, 1998). To address these limitations, Geographically Weighted Models (GWMs) have been introduced to more accurately calibrate both explanatory and predictive models for housing resale prices.

The objective of this take-home exercise is to develop a spatially informed model to predict HDB resale prices in Singapore, making use of appropriate geospatial analytics methods.

We will develop a model to predict HDB resale prices for the period July to September 2025, using HDB resale transaction data from July 2024 to June 2025.

## Installing and Loading R packages

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, Metrics, tidyverse,httr, tidyverse, 
               jsonlite, progress)
```

## Data Preparation

### HDB Dataset

#### Selecting Date of Interest 
This data set comprises of HDB resale transactions from Jan 2017 onwards. For the purpose of the hands on Exercise, the code chunk below will be used to extract a sub-set of transaction records on 5-Room flat transacted from July 2024 to September 2025.

```{r}
HDBresale <- read_csv(
  "data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "5 ROOM") %>%
  filter(month >= "2024-07" & month < "2025-10")
```

#### Geocoding HDB resale data
In this section, we will geocode HDB resale data by using SLA Onemap Geocode API.

Let us prepare a geocoding function by using the code chunk below. It aims to perform the following tasks:

Combining the block and street name into an address. The addresses will be used to geocode instead of postal code that you learned in In-class Exercise 3b,
Passing the address as the searchVal in our query,
Sending the query to OneMapSG search. Note: Since we don’t need all the address details, we can set getAddrDetails as ‘N’,
Converting response (JSON object) to text,
Saving response in text form as a dataframe, and
Retain the latitude and longitude for our output.

```{r}
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address, 
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  res <- GET(base_url, query = query)
  restext<-content(res, as="text")
  
  output <- fromJSON(restext)  %>% 
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)

  return(output)
}
```
Next, the code chunk below will be used to to execute the geocoding function:

```{r}
#| eval: False
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for (i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```
```{r}
#| eval: false
#| include: false
saveRDS(HDBresale, file = "data/rds/HDBresale_geocoded.rds")
```

```{r}
#| include: false
HDBresale <- readRDS("data/rds/HDBresale_geocoded.rds")
```

#### Converting storey_range into a Numerical Variable
### Converting `storey_range` to a Numerical Value (`storey_mid`)

The `storey_range` field in the HDB resale dataset is provided as a text range such as `"07 TO 09"` or `"13 TO 15"`. Since machine learning models require numerical inputs, we convert this range into a single numeric value that approximates the floor level of the unit.

This is done by:

1. **Removing spaces** from the text  
   - Example: `"07 TO 09"` → `"07TO09"`

2. **Splitting the range** into the lower and upper floor values  
   - `"07TO09"` → `"07"` and `"09"`

3. **Calculating the midpoint** (average) of the two values  
   - \[
     \frac{7 + 9}{2} = 8
     \]

4. **Creating a new variable** named `storey_mid` to store this midpoint value.

This numeric representation allows the floor level information to be used effectively in statistical and machine learning models.

5. Remove the original storey_range column, as it is no longer needed.


```{r}
storey_mid <- function(x) {
  x %>%
    str_replace_all(" ", "") %>%    # remove spaces
    str_split("TO") %>%             # split into lower and upper floors
    sapply(\(v) mean(as.numeric(v)))  # convert to numeric and take midpoint
}

HDBresale <- HDBresale %>%
  mutate(storey_mid = storey_mid(storey_range)) %>%  # create numeric floor feature
  select(-storey_range)                              # remove the original text column

```

#### Converting `remaining_lease` to a Numeric Value (`remaining_lease_yrs`)

The `remaining_lease` field is provided as text (e.g., `"87 years 4 months"`). To use this information in modelling, we convert it into a numeric value representing the total lease remaining in **years**.

The function `convert_remaining_lease()`:

1. Extracts the number of **years**.
2. Extracts the number of **months** (or uses 0 if not specified).
3. Converts months into a fraction of a year (`months / 12`).
4. Adds the two together to get a single numeric value.

We then apply the function to create a new column `remaining_lease_yrs` and drop the original text column.

```{r}
convert_remaining_lease <- function(x) {
  years <- as.numeric(str_extract(x, "\\d+(?=\\syear)"))
  months <- str_extract(x, "(?<=years?\\s)\\d+(?=\\smonth)") %>% as.numeric()
  months[is.na(months)] <- 0
  years + months/12
}

HDBresale <- HDBresale %>%
  mutate(remaining_lease_yrs = convert_remaining_lease(remaining_lease)) %>%
  select(-remaining_lease)
```

### Proximity Analysis
In this section, we will learn how to count the number of geographic entities located within a defined distance from each HDB property.

#### Converting to an sf object
Before performing proximity analysis, it is important to ensure that:

- input data sets must be in sf objects, and
- they must be in similar projected coordinates systems.

In the code chunk below,

- st_as_sf() is used to convert HDBresale tibble data.frame to sf object by using values from LONGITUDE and LATITUDE columns to form the geometry features.
- st_transform() is then used to transform the sf object into svy21 projected coordinates system.
- the output sf object is called HDBresale_sf. It is in sf data.frame format.


```{r}
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326) %>%
  st_transform(crs = 3414)
```
#### PreSchool Location 
Next, we will locate and download Pre-Schools Location data from Singapore’s open data portal. There are both geojson and kml version. In this exercise, the kml version will be used.

Then, code chunk below will be used to import the preschool location data into R environment by using st_read() of readr package.

```{r}
preschool = st_read(
  "data/rawdata/PreSchoolsLocation.kml") %>%
  st_transform(crs= 3414)
```
Lastly, code chunk below will be used to count the number of pre-schools located within 350m of each HDB property.

```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(WITHIN_350M_PRESCHOOL = map_int(within_350m, length))
```



