---
title: "Take Home Exercise 1"
author: "Gautamgovan"
date: "September 13, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    number-sections: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview 

This take-home exercise investigates the **spatial** (“where”) and **spatio-temporal** (“where + when”) patterns of **new business registrations in Singapore** between **1 Jan 2024 and 30 Jun 2025**. Using the ACRA “Information on Corporate Entities” datasets and SSIC 2020 classifications, we select at least **three business types** (2- or 3-digit SSIC) and analyse how their locations and intensities evolve over time.

Our workflow mirrors the point-pattern framework taught in class:

- **Data wrangling & geocoding:** tidy ACRA records, standardise dates/SSIC codes, and geocode by postal code via OneMap; convert to `sf` with SVY21 (EPSG:3414).
- **Exploratory mapping:** counts by subzone/planning area to establish baseline patterns.
- **First-order analysis (intensity):** **Kernel Density Estimation (KDE)** to reveal spatial hot spots for each SSIC and period.
- **Second-order analysis (dependence):** **Ripley’s K / CSR envelopes** to test whether registrations are clustered, random, or dispersed.
- **Spatio-temporal analysis:** **STKDE** (monthly slices) to detect **emerging, persistent, and fading** clusters over the study window.

The results are communicated through cartographic maps and concise figure captions  words , followed by a synthesis translating findings into **urban land-use and economic development implications**—e.g., infrastructure readiness, zoning flexibility, activation of under-served subzones, and monitoring of over-concentration risks.

## The Data

This study uses ACRA corporate entity registrations combined with reference layers and services to enable spatial and spatio-temporal analysis.

### Data sources

| Dataset Name | Description | Format | Provider | Notes |
|---|---|---|---|---|
| **ACRA – Information on Corporate Entities (A–Z, Others)** | All entity registrations. Used to extract new businesses by SSIC and date. | CSV (27 files) | data.gov.sg / ACRA | Split alphabetically: A–Z plus a 27th file for names starting with digits/symbols. |
| **SSIC 2020** | Singapore Standard Industrial Classification; used to choose business types (2-/3-/5-digit). | CSV/XLSX | SingStat | We derive `ssic2`, `ssic3` and use `primary_ssic_code`. |
| **OneMap Search API** | Geocoding service to convert postal codes to coordinates. | Web API | SLA / OneMap | Returns WGS84 (lat/lon). We transform to SVY21 (EPSG:3414). |
| **Master Plan Subzones (MPSZ 2024)** | Administrative subzone polygons for aggregation and mapping. | GPKG/Shapefile | data.gov.sg / URA/SLA | Reprojected to SVY21 (EPSG:3414) for analysis. |

### Fields used (ACRA)

| Field | Meaning | Example |
|---|---|---|
| `uen` | Unique Entity Number (entity identifier) | 201912345K |
| `entity_name` | Registered business name | ABC PTE. LTD. |
| `registration_incorporation_date` | Registration date (YYYY-MM-DD) | 2025-03-14 |
| `primary_ssic_code` | Primary SSIC code (2/3/4/5-digit) | 56111 |
| `postal_code` | Registered postal code (should be 6 digits) | 018989 |
| `entity_type` / `status` | Type/status if required for filtering | LIVE COMPANY |

> Additional columns may exist in different ACRA dumps; we standardise names with `janitor::clean_names()` and coalesce equivalent fields where needed.

### Study coverage

- **Temporal window:** 1 Jan 2024 – 30 Jun 2025 (inclusive).  
- **Spatial extent:** Singapore national boundary; mapping/aggregation by **subzone** and **planning area**.  
- **Business scope:** At least **three** SSIC types (2- or 3-digit groups, or 5-digit subclasses), selected prior to analysis.

### Pre-processing & wrangling (summary)

- Read and row-bind **all 27 ACRA CSVs**; add `source_file` for provenance.  
- Standardise fields: parse `registration_incorporation_date` as Date (UTC), derive `year`, `month`.  
- **Postal code hygiene:** keep digits only, left-pad to 6; drop non-6-digit values before geocoding.  
- **Deduplicate** by `uen` + `registration_incorporation_date` (if necessary).  
- Filter to the target SSIC code set and to the study window.  
- **Geocode** unique postal codes via OneMap; join `X`/`Y` back to ACRA records.  
- Convert to `sf` (WGS84) → transform to **SVY21 (EPSG:3414)** for distance-based methods.  
- Join subzone attributes; save intermediates as RDS for reproducibility.

### Data quality & limitations

- **Geocoding precision:** Postal codes resolve to building/entrance centroids, not unit-level; some industrial sites share codes.  
- **Registration vs. operation:** Registration date ≠ opening date; counts reflect legal registrations.  
- **Address anomalies:** Legacy/PO box or malformed postcodes may not geocode; these are excluded.  
- **Classification drift:** SSIC codes can change over time; we use the recorded `primary_ssic_code` as-is.

### Coordinate reference systems (CRS)

- **Input:** WGS84 (EPSG:4326) from OneMap.  
- **Analysis/Maps:** SVY21 / Singapore TM (EPSG:3414), metres, suitable for KDE and K-functions.

### Ethical/usage notes

- ACRA data contains **entity names** (public corporate info), not personal data.  
- Results are aggregated for planning insights; avoid singling out individual firms.

## Load Packages

Following code chunk is used to load neccessary packages 

```{r}
pacman::p_load(
  tidyverse,   # dplyr/readr/purrr/stringr
  sf, tmap,    # geospatial + mapping
  httr,        # OneMap API calls
  lubridate,   # year(), month(), etc.
  spatstat,    # KDE, K-function, CSR (monolithic ok)
  spNetwork,   # network KDE (optional/bonus)
  sparr,       # relative risk (optional/bonus)
  plotly,      # interactive plots (optional)
  gtsummary,   # summary tables 
  dplyr,
  janitor,
  stringr
)
```
## Data Wrangling 
### Preparing the Subzone Boundary (why & how)

This chunk converts URA’s **Master Plan Subzone (No Sea)** GeoJSON into a clean, projected **GeoPackage** with consistent field names. We use it as:
1) the **analysis window** for point-pattern methods, and  
2) the **join layer** to aggregate new businesses by subzone/planning area.
Extract data for planning 


```{r}
sz_raw_path <- "data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson"

subzones <- st_read(sz_raw_path, quiet = TRUE) %>%
  clean_names() %>%                 # lower_case_names
  st_make_valid() %>%               # fix any tiny geometry issues
  st_transform(3414)                # SVY21 metres (needed for KDE/K-functions)
```

Standardise key attribute names 
```{r}
# Helper: extract a field value from the HTML in `description`
extract_field <- function(html, key){
  m <- str_match(html, paste0("<th>", key, "</th>\\s*<td>([^<]*)</td>"))
  val <- m[, 2]
  val <- ifelse(is.na(val) | val == "", NA_character_, str_squish(val))
  return(val)
}

subzones_std <- subzones %>%
  mutate(
    SUBZONE_NO = as.integer(extract_field(description, "SUBZONE_NO")),
    SUBZONE_N  = extract_field(description, "SUBZONE_N"),
    SUBZONE_C  = extract_field(description, "SUBZONE_C"),
    CA_IND     = extract_field(description, "CA_IND"),
    PLN_AREA_N = extract_field(description, "PLN_AREA_N"),
    PLN_AREA_C = extract_field(description, "PLN_AREA_C"),
    REGION_N   = extract_field(description, "REGION_N"),
    REGION_C   = extract_field(description, "REGION_C"),
    FMEL_UPD_D = extract_field(description, "FMEL_UPD_D")
  ) %>%
  select(SUBZONE_NO, SUBZONE_N, SUBZONE_C, CA_IND,
         PLN_AREA_N, PLN_AREA_C, REGION_N, REGION_C, FMEL_UPD_D, geometry) %>%
  st_make_valid() %>%
  st_transform(3414)

# save a clean, fast-to-load copy
st_write(subzones_std, "data/geospatial/mpsz_subzones_2019_nosea.gpkg",
         layer = "subzones", delete_dsn = TRUE)
```




