---
title: "Take Home Exercise 1"
author: "Gautamgovan"
date: "September 13, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    number-sections: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview 

This take-home exercise investigates the **spatial** (“where”) and **spatio-temporal** (“where + when”) patterns of **new business registrations in Singapore** between **1 Jan 2024 and 30 Jun 2025**. Using the ACRA “Information on Corporate Entities” datasets and SSIC 2020 classifications, we select at least **three business types** (2- or 3-digit SSIC) and analyse how their locations and intensities evolve over time.

Our workflow mirrors the point-pattern framework taught in class:

- **Data wrangling & geocoding:** tidy ACRA records, standardise dates/SSIC codes, and geocode by postal code via OneMap; convert to `sf` with SVY21 (EPSG:3414).
- **Exploratory mapping:** counts by subzone/planning area to establish baseline patterns.
- **First-order analysis (intensity):** **Kernel Density Estimation (KDE)** to reveal spatial hot spots for each SSIC and period.
- **Second-order analysis (dependence):** **Ripley’s K / CSR envelopes** to test whether registrations are clustered, random, or dispersed.
- **Spatio-temporal analysis:** **STKDE** (monthly slices) to detect **emerging, persistent, and fading** clusters over the study window.

The results are communicated through cartographic maps and concise figure captions  words , followed by a synthesis translating findings into **urban land-use and economic development implications**—e.g., infrastructure readiness, zoning flexibility, activation of under-served subzones, and monitoring of over-concentration risks.

## The Data

This study uses ACRA corporate entity registrations combined with reference layers and services to enable spatial and spatio-temporal analysis.

### Data sources

| Dataset Name | Description | Format | Provider | Notes |
|---|---|---|---|---|
| **ACRA – Information on Corporate Entities (A–Z, Others)** | All entity registrations. Used to extract new businesses by SSIC and date. | CSV (27 files) | data.gov.sg / ACRA | Split alphabetically: A–Z plus a 27th file for names starting with digits/symbols. |
| **SSIC 2020** | Singapore Standard Industrial Classification; used to choose business types (2-/3-/5-digit). | CSV/XLSX | SingStat | We derive `ssic2`, `ssic3` and use `primary_ssic_code`. |
| **OneMap Search API** | Geocoding service to convert postal codes to coordinates. | Web API | SLA / OneMap | Returns WGS84 (lat/lon). We transform to SVY21 (EPSG:3414). |
| **Master Plan Subzones (MPSZ 2024)** | Administrative subzone polygons for aggregation and mapping. | GPKG/Shapefile | data.gov.sg / URA/SLA | Reprojected to SVY21 (EPSG:3414) for analysis. |

### Fields used (ACRA)

| Field | Meaning | Example |
|---|---|---|
| `uen` | Unique Entity Number (entity identifier) | 201912345K |
| `entity_name` | Registered business name | ABC PTE. LTD. |
| `registration_incorporation_date` | Registration date (YYYY-MM-DD) | 2025-03-14 |
| `primary_ssic_code` | Primary SSIC code (2/3/4/5-digit) | 56111 |
| `postal_code` | Registered postal code (should be 6 digits) | 018989 |
| `entity_type` / `status` | Type/status if required for filtering | LIVE COMPANY |

> Additional columns may exist in different ACRA dumps; we standardise names with `janitor::clean_names()` and coalesce equivalent fields where needed.

### Study coverage

- **Temporal window:** 1 Jan 2024 – 30 Jun 2025 (inclusive).  
- **Spatial extent:** Singapore national boundary; mapping/aggregation by **subzone** and **planning area**.  
- **Business scope:** At least **three** SSIC types (2- or 3-digit groups, or 5-digit subclasses), selected prior to analysis.

### Pre-processing & wrangling (summary)

- Read and row-bind **all 27 ACRA CSVs**; add `source_file` for provenance.  
- Standardise fields: parse `registration_incorporation_date` as Date (UTC), derive `year`, `month`.  
- **Postal code hygiene:** keep digits only, left-pad to 6; drop non-6-digit values before geocoding.  
- **Deduplicate** by `uen` + `registration_incorporation_date` (if necessary).  
- Filter to the target SSIC code set and to the study window.  
- **Geocode** unique postal codes via OneMap; join `X`/`Y` back to ACRA records.  
- Convert to `sf` (WGS84) → transform to **SVY21 (EPSG:3414)** for distance-based methods.  
- Join subzone attributes; save intermediates as RDS for reproducibility.

### Data quality & limitations

- **Geocoding precision:** Postal codes resolve to building/entrance centroids, not unit-level; some industrial sites share codes.  
- **Registration vs. operation:** Registration date ≠ opening date; counts reflect legal registrations.  
- **Address anomalies:** Legacy/PO box or malformed postcodes may not geocode; these are excluded.  
- **Classification drift:** SSIC codes can change over time; we use the recorded `primary_ssic_code` as-is.

### Coordinate reference systems (CRS)

- **Input:** WGS84 (EPSG:4326) from OneMap.  
- **Analysis/Maps:** SVY21 / Singapore TM (EPSG:3414), metres, suitable for KDE and K-functions.

### Ethical/usage notes

- ACRA data contains **entity names** (public corporate info), not personal data.  
- Results are aggregated for planning insights; avoid singling out individual firms.

## Load Packages

Following code chunk is used to load neccessary packages 

```{r}
pacman::p_load(
  tidyverse,   # dplyr/readr/purrr/stringr
  sf, tmap,    # geospatial + mapping
  httr,        # OneMap API calls
  lubridate,   # year(), month(), etc.
  spatstat,    # KDE, K-function, CSR (monolithic ok)
  spNetwork,   # network KDE (optional/bonus)
  sparr,       # relative risk (optional/bonus)
  plotly,      # interactive plots (optional)
  gtsummary,   # summary tables 
  dplyr,
  janitor,
  stringr,
  readr,
  purrr
)
```
## Data Wrangling 
### Preparing the Subzone Boundary (why & how)

This chunk converts URA’s **Master Plan Subzone (No Sea)** GeoJSON into a clean, projected **GeoPackage** with consistent field names. We use it as:
1) the **analysis window** for point-pattern methods, and  
2) the **join layer** to aggregate new businesses by subzone/planning area.
Extract data for planning 


```{r}
sz_raw_path <- "data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson"

subzones <- st_read(sz_raw_path, quiet = TRUE) %>%
  clean_names() %>%                 # lower_case_names
  st_make_valid() %>%               # fix any tiny geometry issues
  st_transform(3414)                # SVY21 metres (needed for KDE/K-functions)
```

Standardise key attribute names 
```{r}
# Helper: extract a field value from the HTML in `description`
extract_field <- function(html, key){
  m <- str_match(html, paste0("<th>", key, "</th>\\s*<td>([^<]*)</td>"))
  val <- m[, 2]
  val <- ifelse(is.na(val) | val == "", NA_character_, str_squish(val))
  return(val)
}

subzones_std <- subzones %>%
  mutate(
    SUBZONE_NO = as.integer(extract_field(description, "SUBZONE_NO")),
    SUBZONE_N  = extract_field(description, "SUBZONE_N"),
    SUBZONE_C  = extract_field(description, "SUBZONE_C"),
    CA_IND     = extract_field(description, "CA_IND"),
    PLN_AREA_N = extract_field(description, "PLN_AREA_N"),
    PLN_AREA_C = extract_field(description, "PLN_AREA_C"),
    REGION_N   = extract_field(description, "REGION_N"),
    REGION_C   = extract_field(description, "REGION_C"),
    FMEL_UPD_D = extract_field(description, "FMEL_UPD_D")
  ) %>%
  select(SUBZONE_NO, SUBZONE_N, SUBZONE_C, CA_IND,
         PLN_AREA_N, PLN_AREA_C, REGION_N, REGION_C, FMEL_UPD_D, geometry) %>%
  st_make_valid() %>%
  st_transform(3414)

# save a clean, fast-to-load copy
st_write(subzones_std, "data/geospatial/mpsz_subzones_2019_nosea.gpkg",
         layer = "subzones", delete_dsn = TRUE)
```
### Load & Tidy Acra 

```{r}
aspatial_dir <- "data/aspatial"
rds_dir      <- "data/rds"
dir.create(rds_dir, showWarnings = FALSE, recursive = TRUE)
file_list <- list.files(aspatial_dir, pattern = "^ACRA.*\\.csv$", full.names = TRUE)

acra_raw <- file_list |>
  map_dfr(~ read_csv(.x, show_col_types = FALSE)) |>
  clean_names()

# 2) Tidy dates, SSIC, postal codes; derive time fields; filter study window
acra_tidy <- acra_raw |>
  mutate(
    reg_date  = as.Date(registration_incorporation_date),
    year      = year(reg_date),
    month_num = month(reg_date),
    month     = month(reg_date, label = TRUE, abbr = TRUE),

    # keep only digits in postal_code; left-pad to 6
    postal_code = postal_code |>
      str_replace_all("[^0-9]", "") |>
      str_pad(width = 6, side = "left", pad = "0"),

    # normalise primary SSIC to 5 chars, and derive 2-/3-digit prefixes
    primary_ssic_code = str_pad(str_sub(primary_ssic_code, 1, 5), 5, pad = "0"),
    ssic2 = str_sub(primary_ssic_code, 1, 2),
    ssic3 = str_sub(primary_ssic_code, 1, 3)
  ) |>
  filter(!is.na(reg_date),
         reg_date >= as.Date("2024-01-01"),
         reg_date <= as.Date("2025-06-30"))
```
### Select the 3 Industries 
Choosing the three industries

The assignment requires at least three business types. We select broad, policy-relevant **2-digit SSIC** groups that span consumer, knowledge, and logistics activities:

- **56 — Food & Beverage Service Activities** (restaurants, cafés, bars)  
- **62 — Computer programming, IT consultancy & related activities**  
- **52 — Warehousing & support activities for transportation**  

In the code below, we:

1. **Build** a small lookup to attach readable sector labels to each 2-digit SSIC.  
2. **Filter** the cleaned ACRA records to these codes.  
3. **Produce** a quick *sector × year* table to sanity-check sample sizes for 2024 and 2025.  



```{r}
# 56 = Food & Beverage, 62 = Information Technology, 52 = Warehousing & support activities for transport
targets <- tibble(
  ssic2 = c("56", "62", "52"),
  sector = c("F&B (56)", "IT (62)", "Logistics/Warehousing (52)")
)

biz_sel <- acra_tidy |>
  filter(ssic2 %in% targets$ssic2) |>
  left_join(targets, by = "ssic2")

table(biz_sel$sector, biz_sel$year)
```
### Geocode unique postcodes with OneMap 

```{r}
#| label: geo code chunk 
#| echo: true
#| eval: false
# Helper that hits OneMap Search API for one postcode
onemap_lookup <- function(pc){
  res <- GET("https://onemap.gov.sg/api/common/elastic/search",
             query = list(searchVal = pc, returnGeom = "Y", getAddrDetails = "Y", pageNum = "1"))
  js <- content(res)
  if (!is.null(js$found) && js$found > 0) {
    r <- js$results[[1]]
    tibble(POSTAL = r$POSTAL, X = as.numeric(r$X), Y = as.numeric(r$Y))
  } else {
    tibble(POSTAL = pc, X = NA_real_, Y = NA_real_)
  }
}

# Geocode unique postcodes
pcs <- biz_sel$postal_code |> unique()
geo_tbl <- map_dfr(pcs, ~ tryCatch(onemap_lookup(.x), error = function(e)
  tibble(POSTAL = .x, X = NA_real_, Y = NA_real_)))

# Join back and keep geocoded rows
biz_geo <- biz_sel |>
  left_join(geo_tbl, by = c("postal_code" = "POSTAL")) |>
  filter(!is.na(X), !is.na(Y))
```

### Convert to sf, project to SVY21, and join subzones
```{r}
#| label: make-sf-and-join
#| echo: true
#| eval: false
biz_pts <- st_as_sf(biz_geo, coords = c("X","Y"), crs = 4326) |>
  st_transform(3414) |>
  st_join(subzones_std |> dplyr::select(SUBZONE_N, PLN_AREA_N), left = TRUE)

# Save intermediates for fast reload
write_rds(acra_tidy, file.path(rds_dir, "acra_tidy_2024H1_2025H1.rds"))
write_rds(biz_pts,  file.path(rds_dir, "biz_points_geocoded.rds"))
```
### Load precomputed outputs
```{r}
#| label: load-precomputed
#| echo: true
#| eval: true
#| message: false
#| warning: false

acra_tidy <- readr::read_rds(file.path(rds_dir, "acra_tidy_2024H1_2025H1.rds"))
biz_pts   <- readr::read_rds(file.path(rds_dir, "biz_points_geocoded.rds"))

```

### Quick sanity check 
#### counts, CRS, validity
```{r}
#| label: qc-basic
#| message: false
#| warning: false

# Expect these to exist from earlier chunks:
stopifnot(exists("subzones_std"), exists("biz_pts"))

# Basic shapes
nrow(subzones_std); nrow(biz_pts)

# CRS should be SVY21 / EPSG:3414
st_crs(subzones_std)$epsg
st_crs(biz_pts)$epsg

# Any invalid geometries?
all(st_is_valid(subzones_std))
all(st_is_valid(biz_pts))

# Any missing coordinates?
sum(is.na(st_coordinates(biz_pts)[,1]))
sum(is.na(st_coordinates(biz_pts)[,2]))

# Quick distribution check
biz_pts |>
  st_drop_geometry() |>
  count(sector, year, name = "n") |>
  arrange(sector, year)
```
#### Are points inside Singapore subzones?
```{r}
#| label: qc-coverage
#| echo: true
#| eval: false
inside <- st_intersects(biz_pts, subzones_std, sparse = TRUE)
mean(lengths(inside) > 0)   # share of points that fall in a subzone (expect ~1)
```
we find that intersects results in 0 which means some thing went wrong somewhere
therefore we rectify this in the next Sub section 

#### Rebuild the sf points with the correct CRS directly (no transform)
```{r}
#| echo: true
#| eval: false
# X/Y from OneMap are already SVY21 metres → set crs = 3414 and DO NOT transform
biz_pts <- st_as_sf(biz_geo, coords = c("X","Y"), crs = 3414) |>
  st_join(subzones_std |> dplyr::select(SUBZONE_N, PLN_AREA_N), left = TRUE)

# Quick sanity
stopifnot(st_crs(biz_pts)$epsg == 3414)
mean(lengths(st_intersects(biz_pts, subzones_std)) > 0)  # should be ~1

# Overwrite the saved file
write_rds(biz_pts, file.path(rds_dir, "biz_points_geocoded.rds"))
```

#### Load precomputed outputs again
```{r}
#| label: reload_load-precomputed
#| echo: true
#| eval: true
#| message: false
#| warning: false

acra_tidy <- readr::read_rds(file.path(rds_dir, "acra_tidy_2024H1_2025H1.rds"))
biz_pts   <- readr::read_rds(file.path(rds_dir, "biz_points_geocoded.rds"))

```
#### Quick Check again
```{r}

#| label: qc-rebuild-using-biz-pts
#| eval: true
# Coordinate ranges in SVY21 metres
coords <- sf::st_coordinates(biz_pts)
summary(coords)

# Validity + missing checks
all(sf::st_is_valid(biz_pts))
sum(is.na(coords[,1]))
sum(is.na(coords[,2]))

# Coverage inside subzones
inside <- sf::st_intersects(biz_pts, subzones_std, sparse = TRUE)
mean(lengths(inside) > 0)   # should be ~1
```
Now intersects() does leave 1 which means all points are in Singapore

#### Quick Map 
```{r}
tmap_mode("plot")
tm_shape(subzones_std) + tm_polygons(col = "grey90", border.col = "white") +
tm_shape(biz_pts) + tm_dots(size = 0.02, alpha = 0.5)
```
### Removing unwanted planning areas


To avoid empty polygons biasing our maps and point-pattern statistics, we trim the study area to places that actually have registrations. Specifically, we:

- **Keep** planning areas with ≥ 1 geocoded point (computed from `biz_pts`).  
- **Exclude** *NORTH-EASTERN ISLANDS* on purpose (offshore, no relevant activity for our scope).  
- **Keep** *WESTERN ISLANDS* because logistics/port activity is plausible.  
- **Filter** both the polygon layer (`subzones_std → subzones_use`) and the points (`biz_pts → biz_pts_use`) so subsequent joins and KDE/K-functions use a consistent window.  

This reduces noise, speeds processing, and prevents intensity estimates from being diluted by large empty areas.

```{r}
# Planning areas that have at least one point
keep_pa <- biz_pts |>
  st_drop_geometry() |>
  count(PLN_AREA_N, name = "n") |>
  filter(n > 0) |>
  pull(PLN_AREA_N)

subzones_use <- subzones_std |> filter(PLN_AREA_N %in% keep_pa)

subzones_use <- subzones_std |>
  dplyr::filter(PLN_AREA_N %in% keep_pa,
                PLN_AREA_N != "NORTH-EASTERN ISLANDS")

#  also drop any biz points that lie in North-Eastern Islands,
# or that didn’t pick up a planning area for any reason.
biz_pts_use <- biz_pts |>
  dplyr::filter(!is.na(PLN_AREA_N),
                PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

#### Map Check 
```{r}
tmap_mode("plot")
tm_shape(subzones_use) + tm_polygons(col = "grey90", border.col = "white") +
tm_shape(biz_pts) + tm_dots(size = 0.02, alpha = 0.5)
```

