---
title: "In class Exercise 5"
author: "Gautamgovan"
date: "Sep 27, 2025"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 4 
    number-sections: true
    
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true 
---

# In class Exercise 5: Global and Local Measures of Spatial Autocorrelation: sfdep methods

## Load Packages
```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

### The data


```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan") %>% 
  st_transform(crs = 32650)
```

```{r}

hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```
## Global Measures of Spatial Association
### Deriving queens' contiguity
```{r}
wm_q <- hunan %>%
  mutate(
    nb = st_contiguity(geometry),
    wt = st_weights(nb, style = "W"),
    .before = 1
  )
```

### Performing Global Moran'I permutation test

```{r}
set.seed(1234)
global_moran_perm(wm_q$GDPPC,wm_q$nb,wm_q$wt,nsim=99)
```
Reject null hypothesis as p value less than alpha value of 0.05

They are spatially correlated 

In conclusion the GDP per capita per county level shows  spatial clustering 

## Local Moran's I 

```{r}
set.seed(1234)
lisa <- wm_q %>% 
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

```{r}
glimpse(lisa)
```
## Visualising Lisa map 
```{r}
lisa <- lisa %>%
  mutate(
    LISA_cluster = ifelse(
      p_ii_sim < 0.05,
      as.character(mean), 
      "Insignificant"),
    LISA_cluster = factor(
      LISA_cluster,
      levels = c("Insignificant", "Low-Low", "Low-High", "High-Low", "High-High")
    )
  )
```

```{r}
lisa_map <- tm_shape(lisa) + 
  tm_polygons(
    fill = "LISA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",      # Insignificant
        "blue",        # Low-Low
        "lightblue",   # Low-High
        "pink",        # High-Low
        "red"          # High-High
      )
    ),
    fill.legend = tm_legend(
      title = "LISA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("Local Moran's I Clusters (p < 0.05)")
lisa_map
```

# Emerging Hot spot Analysis 



